---
title: "Transfer annotations"
author: "Martin loza"
date: "2023/03/17"
---

In previous analyses I annotated the regions using the main histones and the distance to the nearest gene. Now, I would like to transfer the annotations to the whole dataset

```{r setup}
library(dplyr)
library(ggplot2)
library(patchwork)
library(stringr)

## Global variables
date = "230317"
seed = 777
in_dir = "Data/annotated_filtered//loop_2/"
out_dir = "Data/annotated/" 

## Global functions
source("Analysis/Functions.R")
source("Analysis/Functions_Visualizations.r")

#Local functions

```

## Load data

Load annotated regions

```{r}
regions_ann <- readRDS(file = paste0(in_dir, "/u_regions_NS_annotated_230306.rds"))

#short number of regions  
test <- table(regions_ann$type)[c("CTS", "NS", "HK")] == c(27737,300,10965)
if((!test[1]) | (!test[2])){
  cat("Number of regions do not match!!!!")
}else{
  cat("Number of NS, CTS, and HK regions OK")
}
rm(test)
```

Load regions 

```{r}
regions <- readRDS(file = "Data/merged/merged_nearest_gene_221130.rds")
```

## Set up

Let's remove the negative samples

```{r}
regions_ann <- regions_ann %>% filter(type != "NS")
```

During the annotation process we filtered-out suspicious regions that overlapped with NS and presented high levels of Ns 

Let's filter these regions from the original dataset too

```{r}
keep_idx <- which(regions$region_id %in% regions_ann$region_id)
regions <- regions[keep_idx,]
rm(keep_idx)
```

Short test

Number of unique regions 
```{r}
cat("Same number of unique regions? \n") 
nrow(regions %>% filter(!duplicated(region_id))) == nrow(regions_ann)
```

## Transfer the annotations

Let's transfer all the columns that we have in regions_ann but not in regions

```{r}
cols_idx <- which(! colnames(regions_ann) %in% colnames(regions))
sel_cols <- colnames(regions_ann)[cols_idx]
#let's remove columns I don't want to transfer
sel_cols <- sel_cols[-which(sel_cols == "type")]
sel_cols <- sel_cols[-which(sel_cols == "ann_v0")]
sel_cols <- sel_cols[-which(sel_cols == "clusters_louvain_0.2")]
regions <- TransferColumns(reference = regions, query = regions_ann,
                           anchors = "region_id", cols_to_tranfer = sel_cols)
rm(cols_idx, sel_cols)
```

It looks like there is some duplicated informations let's check this

```{r}
identical(regions$dist_to_closest_gene, regions$dist_to_nearest_transcript)
identical(regions$closest_gene, regions$nearest_gene)
identical(regions$closest_gene_id, regions$nearest_transcript_id)
identical(regions$closest_gene_tss, regions$nearest_transcript_tss)
```

Let's remove the duplicated colums

```{r}
sel_columns <- c("dist_to_closest_gene", "closest_gene", "closest_gene_id", 'closest_gene_tss' )
regions <- regions %>% RemoveColumn(column = sel_columns) 
```


```{r}
colnames(regions)
```

## Save data

```{r}
saveRDS(object = regions, file = paste0(out_dir, "/all_regions_annotated_",date,".rds") )
```







