---
title: "Merged results"
author: "Martin loza"
date: "2022/11/30"
---

On this notebook, I want to merge the results of merging overlapping CTSR, HKR, and others.

```{r setup}
library(dplyr)

## Global variables
date = "221130"
seed = 777
in_dir =  "Data/merged"
out_dir = in_dir
```

## Load enhancers data 

Load merged CTSR, HKR, and others

```{r}
#load cell type-specific CREs
ctsr <- readRDS(file = paste0(in_dir, "/ctse_expanded_221126.rds"))
#load HK-CREs
hkr <- readRDS(file = paste0(in_dir, "/hke_expanded_221127.rds"))
#load others (merged CREs from 44 to 2 cell-types)
others <- readRDS(file = paste0(in_dir, "/me_expanded_2_221127.rds"))
```

Load raw data after merging overlapping regions

```{r}
raw_filtered <- readRDS(file = paste0(in_dir, "/raw_filtered_2_221127.rds"))
```

## Analysis of raw_filtered

We expect no regions after the iterative process

```{r}
cat("Number of raw regions after the iterative process: \n")
nrow(raw_filtered)
```

Nice, we don't have any raw regions. We confirmed that the iterative process was run succesfully. We can delete this data frame

```{r}
rm(raw_filtered)
```

## Set up merged regions

Let's confirmed that we have the same number of columns

```{r}
ncol(ctsr) == ncol(hkr)
ncol(hkr) == ncol(others)
```

The CTSR have a unique identifier as raw_CTSE. Let'confirm that only these regions were assigned with this id

```{r}
table(ctsr$raw_CTSE)
table(hkr$raw_CTSE)
table(others$raw_CTSE)
```

The others regions haven't been assigned with this id. Let's assign it

```{r}
others$raw_CTSE <- FALSE
#also assign to hkr just go be sure
hkr$raw_CTSE <- FALSE
```

Let's check the names of the columns

```{r}
sum(colnames(ctsr) == colnames(hkr)) == ncol(ctsr)
sum(colnames(hkr) == colnames(others)) == ncol(hkr)
```

Nice, everything looks good.

## Merge CTSR, HKR, and others

```{r}
regions <- Reduce(f = rbind, x = list(ctsr, hkr, others))
table(regions$test_n_celltypes)
```

```{r}
colnames(regions)
```

## Save merged regions

```{r}
saveRDS(object = regions, file = paste0(out_dir, "/regions_expanded_merged_", date,".rds"))
```














