---
title: "QC. Filter short regions"
author: "Martin loza"
date: "2022/11/30"
---

I hybrid-merged the enhancers and now I would like to filter short length regions.

```{r setup}
library(dplyr)
library(ggplot2)

## Global variables
date = "221130"
seed = 777
in_dir = "Data/merged/"
out_dir = in_dir

# Global functions
source("Analysis/Functions.R")
```

## Load merged enhancers data 

```{r}
regions <- readRDS(file = paste0(in_dir,"/regions_expanded_merged_221130.rds"))
```

Check enhancers

```{r}
CheckEnhancers(enhancers =  regions)
```

## Set up

This objects contains the expanded regions, e.g. one row for each enhancer-gene interaction. Then, we need to get the unique regions 

```{r}
u_regions <- regions %>% filter(!duplicated(region_id))
table(u_regions$test_n_celltypes)
```

## QC

### Regions with number of cell types mismatch

During the merging process there were some regions with a mismatch on the number of cell types. This could be a mismatch between the used packages. For example, we use the bedr to get the core regions, and then transfer the meta data using the GenomicRanges package.
 
Let's get the difference in number of cell types
```{r}
u_regions <- u_regions %>% mutate(diff_n_celltypes = test_n_celltypes - n_celltypes) 
```

```{r}
cat("Number of cells' mismatch: \n")
u_regions %>% filter(n_celltypes < 50) %>% pull(diff_n_celltypes) %>% table()
```

Most of the mismatch regions has a mismatch of one cell types. 
Let's see their distribution by number of cell types

```{r}
u_regions %>% filter(diff_n_celltypes > 0) %>% ggplot(mapping = aes(x = n_celltypes)) + geom_histogram(stat = "count") + theme_bw()
```

Looks like the less number of cell types the more chance to get a mismatch, I think this is kind of expected as regions with less mismatch would be noisier.

If we filtered these regions, how many regions would we filter?

```{r}
cat("Percentage of unique regions with mistmatch:\n")
paste0(100*round(sum(u_regions$diff_n_celltypes > 0)/nrow(u_regions), digits = 5), "%")
```

These regions only represents less than 0.5% of the total regions, and they areless likely to be in HKR. Then, I think we can just filter them to have a stable set of enhancers.

```{r}
#get the ids of mismatched regions
rmv_id <- u_regions %>% filter(diff_n_celltypes > 0) %>% pull(region_id) 
#filter mismtached regions in the expanded data
regions_filtered <- regions %>% filter(!region_id %in% rmv_id)
#short test
sum(regions_filtered$region_id %in% rmv_id) == 0
```

### Filter short regions

Let's update the u_regions 

```{r}
u_regions <- regions_filtered %>% filter(!duplicated(region_id))
table(u_regions$n_celltypes)
```

Short regions can be created during the merging process. I would like to filter these regions using a threshold.

Let's check the summary of length first

```{r}
summary(u_regions$length)
```

There are regions with 0bps length. 

Let's check the histogram of the length

```{r}
u_regions %>% ggplot(mapping = aes(x = length)) + geom_histogram(bins = 80) + scale_x_continuous(trans = "log", breaks = c(50,300,500, 1e3, 1e4))
```

Most of the regions are larger than 390 bps. 

Let's check the length distribution by number of cell types

```{r fig.width=4}
u_regions %>% mutate(n_celltypes = factor(u_regions$n_celltypes)) %>% filter(length <= 2500 ) %>% 
  group_by(n_celltypes) %>% ggplot(mapping = aes(x = n_celltypes, y = length)) +
  geom_boxplot(outlier.size = 0) + geom_point(size = 0.01, alpha = 0.2) + theme_bw() + scale_y_continuous(trans = "sqrt", breaks = c(50,100, 200, 300,500, 1000, 2000))
```

Let's check the percentage of enhancer with different filter thresholds.

```{r}
for(short in c(0,50,100,200)){
  info = paste(paste0("\nNumber of regions: ",nrow(u_regions)), 
                   paste0("Number of short regions: ",sum(u_regions$length <= short)),
                   paste0("Percentage of short regions: ",
                          round(100*sum(u_regions$length <= short)/nrow(u_regions), digits = 2), "%"), 
                   sep = "\n")
  
  cat("\n\nFiltering ", short, "bps short regions. " , info)
}
rm(short, info)
```

If we filter regions shorter than 50bps, we will lose 8887 regions which represents the 3.8%. Also this filter don't interrupt the main distribution of the length by number of cell types. I think this is acceptable

Let's check the distribution by number of cell types if we used 50 bps as the minimum length 

After filtering

```{r fig.width=4}
u_regions %>% mutate(n_celltypes = factor(u_regions$n_celltypes)) %>% filter(length <= 2500 & length >=50 ) %>% 
  group_by(n_celltypes) %>% ggplot(mapping = aes(x = n_celltypes, y = length)) +
  geom_boxplot(outlier.size = 0) + geom_point(size = 0.01, alpha = 0.2) + theme_bw() + scale_y_continuous(trans = "sqrt", breaks = c(50,100, 200, 300,500, 1000, 2000))
```

It looks nice. The statistics improved too.

Let's use this filter 

```{r}
regions_filtered <- regions_filtered %>% filter(length >= 50) 
```

Let's update the unique regions 

```{r}
u_regions <- regions_filtered %>% filter(!duplicated(region_id))
summary(u_regions$length)
```


This is the final distribution of enhancers using this filters

```{r}
subtitle = paste(paste0("Number of enhancers: ", nrow(u_regions)), 
                 paste0("Number of CTSR: ", u_regions %>% filter(n_celltypes == 1) %>% count() %>% pull(n) ),
                 paste0("Number of HKR: ", u_regions %>% filter(n_celltypes == 50) %>% count() %>% pull(n) ),
                 sep = "\n")
u_regions %>% filter(length<3500) %>% ggplot(mapping = aes(x = length)) + geom_histogram(bins = 200) + scale_x_continuous(breaks = c(100,500,750, 1000, 1500, 2500)) + theme_bw() + ggtitle(label = "", subtitle = subtitle)
```

```{r, fig.width=4}
u_regions %>% ggplot(mapping = aes(x = test_n_celltypes)) + geom_histogram(stat = "count") + theme_bw() + ggtitle(label = "", subtitle = subtitle)
```


```{r}
min(u_regions$length)
max(u_regions$length)
```

## CTSE vs HKE. Distance to target analysis.

```{r fig.width=3 }
regions_filtered %>% 
  filter(n_celltypes == 1 | n_celltypes == 50) %>% mutate(n_celltypes = factor(n_celltypes)) %>%
  ggplot(mapping = aes(x = distance, color = n_celltypes, fill = n_celltypes)) +
  geom_histogram(bins = 100, alpha = 0.5, position = "dodge") + theme_bw() +scale_x_continuous( trans = "log",breaks = c(1e3,1e4,1e5,1e6,5e6)) + ggtitle("Distance to target genes")
```

```{r}
cat("CTSE - distance: \n")
u_regions %>% filter(n_celltypes ==1) %>% pull(distance) %>% summary()
cat("HKE - distance: \n")
u_regions %>% filter(n_celltypes == 50) %>% pull(distance) %>% summary()
```

## CTSE vs HKE. Number of target genes analysis.

```{r fig.width=3 }
u_regions %>% 
  filter(n_celltypes == 1 | n_celltypes == 50) %>% mutate(n_celltypes = factor(n_celltypes)) %>%
  ggplot(mapping = aes(x = n_genes, color = n_celltypes, fill = n_celltypes)) +
  geom_histogram(bins = 100, alpha = 0.5, position = "dodge") + 
  theme_bw() +scale_y_continuous( trans = "sqrt", breaks = c(0, 1e2,1e3,1e4,1e5,1e6,5e6)) + ggtitle("Number of target genes")
```

```{r}
cat("CTSE - number of target genes: \n")
u_regions %>% filter(n_celltypes ==1) %>% pull(n_genes) %>% summary()
cat("HKE - number of target genes: \n")
u_regions %>% filter(n_celltypes == 50) %>% pull(n_genes) %>% summary()
```

```{r}
saveRDS(object = regions_filtered, paste0(out_dir, "/merged_expanded_filtered_regions_", date, ".rds"))
```
