---
title: "Annotate target genes of the merged regions"
author: "Martin loza"
date: "2022/11/30"
---

In this notebook I aim to find annotate the target genes of the merged regions.

```{r setup}
library(here)
library(dplyr)

## Global variables
date = "221130"
seed = 777
in_dir <- "Data/"
out_dir = in_dir

## Global functions
source("Analysis/Functions.R")

## Local functions 
#Transfer metadata function
TransferMD <- function(target_df = NULL, from_df = NULL, column = NULL, target_anchors = NULL, from_anchors = NULL){
  #set the anchor iformation as rownames
  rownames(from_df) <- from_df[[from_anchors]]
  #fit the target data frame 
  from_df <- from_df[target_df[[target_anchors]],]
  #short tests
  if(any(is.na(from_df[[from_anchors]]))){
    stop("NA founds. Verify the anchors selected.", call. = TRUE)
  }
  if(identical(target_df[[target_anchors]], from_df[[from_anchors]])){
    target_df[[column]] <- from_df[[column]]
  }else{
    stop("Anchors not identical. Verify the anchors selected.", call. = TRUE)
  }
  #returen the data frame with the new meta data
  return(target_df)
}

```

## Load data

Load raw enhancers data 

```{r}
regions <- readRDS(file = paste0(in_dir, "/merged/merged_expanded_filtered_regions_221130.rds"))
```

Load genes universe

```{r}
genes <- readRDS(file = paste0(in_dir,"/hg38/hg38_genes_universe_221121.rds"))
```

## Preliminary analysis

short test
```{r}
any(duplicated(genes$gene_name))
```

No duplicated names. 

```{r}
#get genes in the selected data
regions_genes <- regions %>% filter(!duplicated(gene)) %>% pull(gene)
cat("Number of genes in the universe:\n")
genes %>% filter(!duplicated(gene_name)) %>% nrow()
cat("Number of genes in the selected regions:\n")
length(regions_genes)
```

## Annotate genes

Let's get the genes contained in the regions' targets

```{r}
genes_targets <- genes %>% filter(gene_name %in% regions_genes) 
```

Let's annotate the biotype of target genes
I coded a function for this. 

```{r}
regions <- TransferMD(target_df = regions, from_df = genes_targets, column = "gene_biotype", target_anchors = "gene", from_anchors ="gene_name")
```

```{r}
#get regions' pcg
regions_pcg <- regions %>% filter(!duplicated(gene)) %>% filter(gene_biotype == "protein_coding") %>% pull(gene)
#get universe pcg
pcg <- genes %>% filter(gene_biotype == "protein_coding") %>% pull(gene_name)
cat("Number of missing PCG in the selected data:\n")
sum(!pcg %in% regions_pcg)
cat("Percentage missing:\n")
paste0(100*round(sum(!pcg %in% regions_pcg)/length(pcg),digits = 5), "%")
```

The selected regions covers around the 97% of the pcg in the universe. 

```{r}
rm(pcg, regions_genes, regions_pcg)
```

## Save data

```{r}
saveRDS(object = regions, file = paste0(out_dir,"/merged/merged_expanded_biotype_annotated_",date,".rds"))
```

