---
title: "Define negative samples using SCREEN + ABC enhancers"
author: "Martin loza"
date: "2023/02/10"
---

In previous analyses I got the gaps from the union of ABC enhances and SCREEN regulatory elements. 

On this notebook, I would like to define the negative samples.

```{r setup}
library(dplyr)
library(GenomicRanges)
library(ggplot2)

## Global variables
date = "230210"
seed = 777

in_dir = "Data/negative_samples/"
out_dir <- in_dir

## Global functions
source("Analysis/Functions.R")
source("Analysis/Functions_Visualizations.r")

#Local functions
```

## Load data 

```{r}
gaps <- readRDS(file = paste0(in_dir, "/gaps_filtered_230210.rds") )
head(gaps)
```

## Define negative control samples

I would like the negative control samples to fulfill the next:

- Length = 500 bps (most of the ABC regions are 500 bps)
- The center of the negative controls should be at leas 250bp away from any expanded ABC (abc +- 500bps). 

Then, the NS should be:
250bps (from the center of NS) +
500bps (expanded NS for the histone modification marks) +
500bps (expanded ABC) +
250 extra bases away from any ABC+SCREEN regions, equals

1500 bps on each side.

Then, I would like to get negative samples by sampling gaps with at least 3000bps

### QC of gaps' length

Let's then filter the gaps to be at least 3000 bps length

```{r}
#lets recalculate the length and filter the gaps
gaps_filtered <- gaps %>% 
  mutate(length = end - start) %>%
  select(chr, start, end, length) %>% 
  filter(length >= 3000)
rm(gaps)
```

```{r}
cat("Number of gaps: \n")
nrow(gaps_filtered)
summary(gaps_filtered$length)
cat("\nSummary of gaps' length: \n")
summary(gaps_filtered$length)
gaps_filtered %>% MyHistogram(x = "length") + scale_x_continuous(trans = "log")
```

We can observe that the number of gaps with length higher than 200000 is low

```{r}
gaps_filtered %>% filter(length >= 200000) %>% MyHistogram(x = "length") + scale_x_continuous(trans = "log")
```

I think we can filter these long regions 

```{r}
gaps_filtered <- gaps_filtered %>% filter(length <= 200000)
cat("Summary of gaps' length: \n")
summary(gaps_filtered$length)
gaps_filtered %>% MyHistogram(x = "length") + scale_x_continuous(trans = "log")
```

### Redefine the gaps

Now, I would like to standardize the gaps to be 500 bp from the center

```{r}
#define the center of gaps
gaps_filtered <- gaps_filtered %>% mutate(center = start + floor(length/2))
#standardize gaps
gaps_filtered <- gaps_filtered %>%
  mutate(start_std = center-250, end_std = center+250) %>% 
  mutate(length_std = end_std - start_std)
```

Good let's save this preliminary gaps

```{r}
saveRDS(object = gaps_filtered, file = paste0(out_dir,"/gaps_standard_length_",date,".rds"))
```

```{r}
#load gaps_filtered
# gaps_filtered <- readRDS(file = paste0(out_dir,"/gaps_standard_length_",date,".rds"))
```


### Define negative samples

Let's define 10,000 negative samples, which is similar to the number of HK-regions in the ABC analysis.

```{r}
set.seed(seed)
negative_sampes <- gaps_filtered %>% 
  mutate(chr = factor(as.character(chr), levels = paste0("chr",c(1:22,"X")))) %>%
  sample_n(size = 10000, replace = FALSE) %>%
  mutate(start = start_std, end = end_std) %>%
  mutate(length = end - start) %>%
  select(chr, start, end, length) %>% arrange(chr, start) %>%
  droplevels()
```

Let's see the distribution of negative samples by chromosome

```{r}
cat("Number of gaps: \n")
nrow(negative_sampes)
summary(negative_sampes$length)
cat("\nSummary of gaps' length: \n")
summary(negative_sampes$length)
negative_sampes %>% MyHistogram(x = "chr", stat = "count") + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Let's save the negative samples 

```{r}
saveRDS(object = negative_sampes, 
        file = paste0(out_dir,"/negative_samples_",date,".rds"))
```



























