---
title: "Setup hg38 genes from ensembl"
author: "Martin Loza"
date: "2022/07/14"
---

On this workflow I setup the genes annotation from ENSEMBL. 
The resulting data frame can be found in the figshare folder of our publication.

```{r setup}
library(dplyr)
library(stringr)
library(here)
```

## Load data

Load the ENSEMBL annotations

```{r}
#change/add your Data directory
data <- read.table(file = "Data/Homo_sapiens.GRCh38.107.chr.gtf", sep = "\t", header = FALSE)
data <- as.data.frame(data)
```

## Setup columns

From ensembl information
* First column is the chromosome
* Second column is the source
* Third column is the feature (gene, variation, etc.)
* Forth column is the start
* Fifth column is the end
* Sixth column is a score value
* Seventh column is the strand
* Eigth column is the Frame (I don't know what do they refer by frame)
* Nineth column is the attribute (some extra infor separated by semicolon)

For now, let's rename the columns

```{r}
colnames(data) <- c("chr", "source", "feature", "start", "end", "score", "strand", "frame", "attributes")
```

From test 3 and 4, the `score` and `frame` columns looks like not necessary for our studies. Then, let's remove them from now

```{r}
rmv_idx <- which(colnames(data) %in% c("score", "frame"))
data <- data[,-rmv_idx]
```

Let's keep for now, only genes

```{r}
data <- data %>% filter(feature == "gene")
table(data$feature)
```

Because we know they are only gene, we can remove this column
```{r}
rmv_idx <- which(colnames(data) == "feature")
data <- data[,-rmv_idx]
```

## Analyse attributes

Let's expand the attributes

```{r}
strsplit(x = data$attributes[1], split = ";")
```
We can observe that there are 4 extra information for every gene "gene_id", "gene_version", "gene_name", "gene_source", and "gene_biotype". Let's try to retrieve this information.

```{r}
gene_info <- unlist(strsplit(x = data$attributes[1], split = "; "))
strsplit(gene_info[1], split = " ")

data$attributes[1] %>% strsplit(split = "; ") %>% unlist() %>% strsplit(split = " ")
```

```{r}
#create columns 
data <- data %>% mutate("gene_id" = NA, "gene_version" = NA, "gene_name" = NA, "gene_source" = NA)
#for each row
for(r in 1:nrow(data)){
current_attributes = data$attributes[r] %>% strsplit(split = "; ") %>% unlist() %>% strsplit(split = " ")
  for(a in 1:length(current_attributes)){
    tmp_attributes <- unlist(current_attributes[a])
    data[r,tmp_attributes[1]] <- tmp_attributes[2]
  }
}
```

let's remove unnecesary characters
```{r}
data$gene_biotype <- gsub(x = data$gene_biotype, pattern = ";", replacement = "")
# data %>% filter(gene_biotype == "protein_coding") %>% group_by(chr) %>% count
# unique(data$gene_biotype)
```

For now, let's remove the "version" of the gene, and the original "source" (is duplicated in "gene_source") and the attributes. 

```{r}
rmv_idx <- which(colnames(data) %in% c("gene_source", "source", "attributes"))
data <- data[,-rmv_idx]
```

## Save data

```{r}
saveRDS(object = data, file = "Data/genes_ENSEMBL_hg38.rds")
```

## Tests.

### Test 1. 

What's in V1, V3, and V7.

Looks like V1 contains chromosome information, V3 contains the element information, and V7 contains the strand information.

```{r}
table(data$V1)
table(data$V3)
table(data$V7)
```

### Test 2. 

```{r}
t <- data %>% filter(V3 == "gene")
t %>% group_by(V1) %>% count()
```

### Test 3. scores?
Looks like there is only dots. We can remove it
```{r}
table(data$score)
```

### Test 4. Frame?

I don't know. For now, let's remove it.

```{r}
table(data$frame)
```
