---
title: "Re-normalization of annotated ABC + NS"
author: "Martin loza"
date: "2023/03/06"
---

In previous workflow I filter out regions using sub-clustering analysis. On this workflow I would like to re-normalize the data for further annotation analysis.

This is the second loop in this sub-clustering annotation/QC

```{r setup}
library(dplyr)
library(ggplot2)
library(patchwork)

## Global variables
date = "230306"
seed = 777
in_dir = "Data/annotated_filtered/"
out_dir = "Data/annotated_filtered/loop_2"

## Global functions
source("Analysis/Functions.R")
source("Analysis/Functions_Visualizations.r")

#Local functions

```

## Load regions

Load annotated regions 

```{r}
regions <- readRDS(file = paste0(in_dir, "/u_regions_NS_filtered_unsupervised_230306.rds"))

#short number of regions  
test <- table(regions$type)[c("CTS", "NS", "HK")] == c(27752,300,10965)
if((!test[1]) | (!test[2])){
  cat("Number of regions do not match!!!!")
}else{
  cat("Number of NS, CTS, and HK regions OK")
}
rm(test)
```

```{r}
regions %>% MyHistogram(x = "n_celltypes", stat = "count") + theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

## Analysis of annotations

I would like to analyse how the normalization of features changes their distribution

### Histones features

```{r  fig.width=8}
sel_regions <- regions

sel_ann <- c("H3K27ac", "H3K4me1", "H3K4me3")
fill = NULL
plots <- list()
for(ann in sel_ann){
  plots[[ann]] <- sel_regions %>% MyHistogram(x = ann, fill = fill, position = "dodge") + ggtitle(ann)
}
(plots[[1]] + plots[[2]] + plots[[3]] & scale_y_continuous(trans = "sqrt")) + plot_layout(ncol = 3, guides = "collect")
rm(label, ann,  fill, plots, sel_regions, sel_ann)
```

They look ok. No normalization on this case

## Distance to nearest TSS

I would like to compare with the abs distance
```{r}
tmp <- regions %>% mutate(abs_distance = abs(dist_to_nearest_transcript)) %>% mutate(distance_log = log1p(abs_distance))
```

```{r}
summary(tmp$dist_to_closest_gene)
summary(tmp$abs_distance)
```

```{r fig.width=4}
p1 <- tmp %>% MyHistogram(x = "dist_to_nearest_transcript", n_bins = 200)+ xlim(c(-1e4,1e4))
p2 <- tmp %>% MyHistogram(x = "abs_distance", n_bins = 200) + xlim(c(-200,1e4))
p3 <- tmp %>% MyHistogram(x = "distance_log", n_bins = 500) 

p1 / p2 / p3
rm(p1, p2, p3)
```

Before I checked and it's not trivial to capture the behavior close to TSS with normalization. Then, if for now I will just calculate the log1p of the abs_distance.

## Normalization

Therefore, the next normalization/scaling will be performed

- Histone marks: nothing
- Distance to nearest TSS: abs + log1p

```{r}
#normalize the distance to nearest TSS: abs + log1p
sel_ann <- c("dist_to_nearest_transcript")
for(ann in sel_ann){
  label = paste0(ann, "_norm")
  regions[[label]] <- log1p(abs(regions[[ann]]))
}
rm(sel_ann, ann)
```

## Final distribution of annotations

```{r fig.width=5}
sel_regions <- regions
sel_ann <- c("H3K27ac", "H3K4me1", "H3K4me3")
fill = NULL
plots <- list()

for(ann in sel_ann){
  label <- paste0("Normalized ",ann)
  plots[[ann]] <- sel_regions %>%  MyHistogram(x = ann) + ggtitle(label)
}
(plots[[1]] + plots[[2]] + plots[[3]]  & scale_y_continuous(trans = "sqrt")) + plot_layout(ncol = 3, guides = "collect")
rm(label, ann,  fill, plots, sel_regions, sel_ann)
```

```{r fig.width=4}
sel_regions <- regions
sel_ann <- paste0(c("dist_to_nearest_transcript"), "_norm")
fill = NULL
plots <- list()

for(ann in sel_ann){
  label <- paste0("Normalized ",ann)
  plots[[ann]] <- sel_regions %>%  MyHistogram(x = ann) + ggtitle(label)
}
(plots[[1]] & scale_y_continuous(trans = "sqrt")) + plot_layout(ncol = 2, guides = "collect")
rm(label, ann,  fill, plots, sel_regions, sel_ann)
```

## Save normalized regions

```{r}
saveRDS(object = regions, paste0(out_dir, "/u_regions_NS_filtered_normalized_", date, ".rds"))
```
