---
title: "Setup embeddings"
author: "Martin loza"
date: "2023/03/06"
---

In previous workflows I performed QC based on the percentage of Ns and the number of overlapping cell types for the histone marks.
I normalized the annotations and scaled them after QC using NS and unsupervised analysis.

On this workflow I would like to setup the different embeddings to use in further studies. Calculating embeddings take a long time so it is a good idea to set them up so I don't have to calculated them every time I make a new analysis.

This is loop 2

```{r setup}
library(dplyr)
library(ggplot2)
library(patchwork)
library(umap)
library(bluster)
library(igraph)
library(scales)
library(viridis)
library(stringr)

## Global variables
date = "230306"
seed = 777
in_dir = "Data/annotated_filtered/loop_2"
out_dir = "Data/embeddings/loop_2/"
my_color = "deepskyblue4"

## Global functions
source("Analysis/Functions.R")
source("Analysis/Functions_Visualizations.r")

#Local functions
```

## Load regions

Load annotations: histone, cpg, distance to nearest tss

```{r}
regions <- readRDS(file = paste0(in_dir, "/u_regions_NS_filtered_scaled_230306.rds"))

#short number of regions  
test <- table(regions$type)[c("CTS", "NS", "HK")] == c(27752,300,10965)
if((!test[1]) | (!test[2])){
  cat("Number of regions do not match!!!!")
}else{
  cat("Number of NS, CTS, and HK regions OK")
}
rm(test)
```

```{r}
regions %>% MyHistogram(x = "n_celltypes", stat = "count") + theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

## Set up

Verify the type labels

```{r}
table(regions$type)
```

Let's remove any columns related to PCA or UMAP
```{r}
#remove columns related to PC. Umap columns are also related to UMAP
idx <- which(str_detect(string = colnames(regions), pattern = "PC"))
if(length(idx) != 0){
  regions <- regions[,-idx]
}
idx <- which(str_detect(string = colnames(regions), pattern = "UMAP"))
if(length(idx) != 0){
  regions <- regions[,-idx]
}
colnames(regions)
rm(idx)
```

## Embeddings

I would like to calculate 3 different embeddings:
- Low resolution: H3K27ac, H3K4me1, H3K4me3, and distance to TSS

### Low resolution

Select features

```{r}
features <- c("H3K27ac_scaled", "H3K4me1_scaled", "H3K4me3_scaled",
              "dist_to_nearest_transcript_norm_scaled")
#short tests
cat("Number of features: \n", length(features))
cat("\nAll features in the data frame OK:\n", sum(features %in% colnames(regions)) == length(features) ) 
cat("\nNAs in features OK: \n", any(!is.na(regions[,features])))
```
Get PCA coordinates

```{r}
#get PCA
set.seed(seed)
pca_data <- prcomp(x = regions[,features], center = FALSE, scale. = FALSE, rank. = length(features))
#elbow plot
plot(y = pca_data$sdev, x = 1:length(pca_data$sdev))
#rotation
pca_data$rotation
```

Select the coordinates and rename the columns: "low"(resolution) 

```{r}
pca_embeddings <- as.data.frame(pca_data$x)
colnames(pca_embeddings)
rm(pca_data)
```

Get UMAP coordinates

```{r}
dims = length(features)
set.seed(seed)
umap_data <- umap(d = pca_embeddings[,1:dims], preserve.seed = TRUE, method = 'umap-learn')
#setup umap embeddings
umap_embeddings <- as.data.frame(umap_data$layout)
colnames(umap_embeddings) <- c("UMAP_1", "UMAP_2")
embeddings <- cbind(pca_embeddings, umap_embeddings)
rm(pca_embeddings, umap_embeddings, umap_data, dims)
```

Setup embeddings before saving

```{r}
rownames(embeddings) <- regions$region_id
```

Save low embeddings

```{r}
saveRDS(object = embeddings, file = paste0(out_dir, "/low_embeddings_", date, ".rds"))
```

Let's visualize the low embeddings

```{r}
tmp_df <- cbind(regions, embeddings)
```

```{r fig.width=4, fig.height=4}
p_size = 0.1
p <- PlotGroups(object = tmp_df, group = "type", dim1 = "PC1", dim2 = "PC2",
                point_size = p_size, high_color = my_color,order = TRUE, text_size = 18)

p$NS + p$CTS + p$`2` + p$`5` + p$`10` +
  p$`15` + p$`20` + p$`25` + p$`30` +
  p$`35` + p$`40` + p$HK  +
  plot_layout(nrow = 3) & NoLegend() & coord_fixed() & NoAxis() 
rm(p, p_size)
```

Features plots

```{r fig.width=4}
p_size = 0.1
umap_columns <- c("UMAP_1",  "UMAP_2")
p <- PlotGroups(object = tmp_df, group = "type",
                dim1 = umap_columns[1],
                dim2 = umap_columns[2],
                point_size = p_size, high_color = my_color,order = TRUE, text_size = 18)

p$NS + p$CTS + p$`2` + p$`5` + p$`10` +
  p$`15` + p$`20` + p$`25` + p$`30` +
  p$`35` + p$`40` + p$HK  +
  plot_layout(nrow = 3) & NoLegend() & coord_fixed() & NoAxis() 

rm(umap_columns, p_size)
```

```{r}
p_size = 0.1
tmp <- tmp_df %>% mutate("Dim1" = UMAP_1, "Dim2" = UMAP_2) 
pf <- list()
for(n in features){
  pf[[n]] <- tmp %>% myFeaturePlot(feature = n, order = TRUE, point_size = p_size, text_size = 14, high_color = my_color) + NoAxis() + scale_colour_viridis_c(option = "magma") + ggtitle(n)
}
rm(tmp, p_size)
```

```{r fig.width=8}
p$NS + p$CTS + p$HK +
  pf[[1]] +  pf[[2]] + pf[[3]] +
  pf[[4]] +
  plot_layout(ncol = 3)  & 
  coord_fixed() & NoAxis()
```















