---
title: "Normalization of annotated ABC + NS"
author: "Martin loza"
date: "2023/02/15"
---

In previous workflows I obtained diverse kind of annotations of the extended ABC + NS (e.g. regions +-500bps). The annotations are:
- Main histone modifications
- Nucleotids content and CpG content
- Nearest TSS
- ABC annotations (number of cell types, number of target genes, etc.)
- Conservation score

I also performed QC based on the percentage of Ns and the number of overlapping cell types for the histone marks.

On this workflow I would like to perform QC of the annotations that need it.

```{r setup}
library(dplyr)
library(ggplot2)
library(patchwork)

## Global variables
date = "230215"
seed = 777
in_dir = "Data/annotated/"
out_dir = in_dir

## Global functions
source("Analysis/Functions.R")
source("Analysis/Functions_Visualizations.r")

#Local functions

```

## Load regions

Load annotated regions 

```{r}
regions <- readRDS(file = paste0(in_dir, "/u_regions_NS_annotated_230215.rds"))

#short number of regions  
test <- table(regions$type) == c(184913,6605)
if((!test[1]) | (!test[2])){
  cat("Number of regions do not match!!!!")
}else{
  cat("Number of NS + ABC regions OK")
}
rm(test)
```

```{r}
colnames(regions)
```

## Normalization

We analized each features independently and how different normalization affect their distribution. For more info check the Github repository containing the whole set of test/results/etc.

The next normalization/scaling will be performed

- Histone marks: nothing
- Nucleotide content (per_AT, per_GC, exp/obs CpG, per_N):  log1p
- Distance to nearest TSS: abs + log1p
- Conservation score" add offset (minimum of the conservation score) + log1p

```{r}
#normalize nucleotide content:  log1p + scaling
sel_ann <- c("pct_GC", "pct_AT", "per_N", "obsExp_CpG")
for(ann in sel_ann){
  label = paste0(ann, "_norm")
  regions[[label]] <- log1p(regions[[ann]])
}
rm(sel_ann, ann)
```

```{r}
#normalize the distance to nearest TSS: abs + log1p
sel_ann <- c("dist_to_nearest_transcript")
for(ann in sel_ann){
  label = paste0(ann, "_norm")
  regions[[label]] <- log1p(abs(regions[[ann]]))
}
rm(sel_ann, ann)
```

```{r}
#normalize the conservation score: add offset (abs of the minimum of the conservation score) + log1p
sel_ann <- c("conservation_score")
for(ann in sel_ann){
  label = paste0(ann, "_norm")
  min_cs <- abs(min(regions$conservation_score))
  regions[[label]] <- log1p(regions[[ann]] + min_cs)
}
rm(sel_ann, ann)
```

## Save normalized regions

```{r}
saveRDS(object = regions, paste0(out_dir, "/u_regions_NS_normalized_", date, ".rds"))
```













