---
title: "Unsupervised analysis"
author: "Martin loza"
date: "2023/02/15"
---

In previous workflows I obtained diverse kind of annotations of the extended ABC + NS (e.g. regions +-500bps). The annotations are:
- Histone modifications
- Nucleotids content and CpG content
- Nearest TSS
- ABC annotations (number of cell types, number of target genes, etc.)
- Conservation score

I performed QC based on the percentage of Ns and the number of overlapping cell types for the histone marks.

I normalized the annotations. Each annotations was independently analysed and the type of normalization were chose accordingly.

On this workflow I would like to perform the first unsupervised analysis of the regions. Most probably I will filter out clusters of regions correlated with NS. So I expect this workflow to serve as an extra QC step.

```{r setup}
library(dplyr)
library(ggplot2)
library(patchwork)
library(umap)
library(bluster)
library(igraph)
library(reticulate)

## Global variables
date = "230215"
seed = 777
in_dir = "Data/annotated/"
out_dir = in_dir
my_color = "deepskyblue4"

## Global functions
source("Analysis/Functions.R")
source("Analysis/Functions_Visualizations.r")

#Local functions

ClusterLouvain <- function(x, k = 10, resolution = 0.5) {
  
  g <- bluster::makeSNNGraph(x, k = k)
  res <- igraph::cluster_louvain(g, resolution = resolution)

  memberships <- igraph::membership(res)
  
  return(memberships)
}

LegendDotSize <- function(size = 5){
  return(guides(colour = guide_legend(override.aes = list(size=size))))
}
```

## Load regions

Load annotations: histone, cpg, distance to nearest tss

```{r}
regions <- readRDS(file = paste0(in_dir, "u_regions_NS_normalized_230215.rds"))

#short number of regions  
test <- table(regions$type) == c(184913,6605)
if((!test[1]) | (!test[2])){
  cat("Number of regions do not match!!!!")
}else{
  cat("Number of NS + ABC regions OK")
}
rm(test)
```

```{r}
colnames(regions)
```

## Set up

Selected features for unsupervised analysis

```{r}
features <- c("H3K27ac"  , "H3K4me1", "H3K4me3",
              "H3K27me3", "H3K36me3", "H3K9ac",
              "H3K9me3",
              "pct_GC_norm", "pct_AT_norm", "per_N_norm", "obsExp_CpG_norm",
              "dist_to_nearest_transcript_norm",
              "conservation_score_norm")

cat("Number of features: \n", length(features))
cat("\nNAs in features OK: \n", any(!is.na(regions[,features])))
```

Add label for visualizations

```{r}
regions <- regions %>% mutate(type = "other") 
types <- c(1,2,5,10,15,20, 25, 30, 35, 40, 50, "NS")
for(t in types){
  #get indexed cells
  idx <- which(regions$n_celltypes == t)
  #set cts type
  if(t == "1"){
    regions$type[idx] <- "CTS"
  }else if(t == "50"){
    regions$type[idx] <- "HK"
  }else{
    regions$type[idx] <- t
  }
}
rm(idx)

regions$type <- factor(regions$type, levels = c("CTS",2,5,10,15,20, 25, 30, 35, 40, "HK", "NS", "other"))

table(regions$type)
rm(types, t)
```

## Unsupervised analysis

## Analysis using the main histone marks

Selected features for unsupervised analysis

```{r}
features <- c("H3K27ac"  , "H3K4me1", "H3K4me3",
              "pct_GC_norm", "pct_AT_norm", "obsExp_CpG_norm",
              "dist_to_nearest_transcript_norm")

cat("Number of features: \n", length(features))
cat("\nNAs in features OK: \n", any(!is.na(regions[,features])))
```

### Scaled data

I would like to save the scaled data separately for future analyses.

```{r}
scaled_data <- apply(X = regions[,features], MARGIN = 2, FUN = scale, center = TRUE, scale = TRUE)
colnames(scaled_data) <- paste0(colnames(scaled_data),"_scaled")
```

### PCA

Let's calcualte the PCA and confirm the variance captured.
```{r}
set.seed(seed)
pca_data <- prcomp(x = scaled_data, center = FALSE, scale. = FALSE, rank. = length(features))
plot(y = pca_data$sdev, x = 1:length(pca_data$sdev))
```

Analyse the relation between PCs and features

```{r}
pca_data$rotation
```

Update embedding data frame

```{r}
data_embeddings <- cbind(regions, scaled_data, as.data.frame(pca_data$x))
```

Visualize the CREs in the first 3 PCs

```{r fig.width=4}
p_size = 0.5
sel_regions <- data_embeddings %>%
  mutate(n_celltypes = factor(n_celltypes)) %>% arrange(n_celltypes)
p1 <- sel_regions %>% ggplot(mapping = aes(x = PC1, y = PC2)) + geom_point(size = p_size, colour = my_color) + theme_bw() + coord_fixed()
p2 <- sel_regions %>% ggplot(mapping = aes(x = PC1, y = PC3)) + geom_point(size = p_size, colour = my_color) + theme_bw() 
p3 <- sel_regions %>% ggplot(mapping = aes(x = PC2, y = PC3)) + geom_point(size = p_size, colour = my_color) + theme_bw() 

p1 + p2 + p3 + plot_layout(guides = "collect") 

rm(sel_regions,p1,p2,p3)
```

```{r fig.width=4, fig.height=4}
p_size = 0.1
p <- PlotGroups(object = data_embeddings, group = "type", dim1 = "PC1", dim2 = "PC2",
                point_size = p_size, high_color = my_color,order = TRUE, text_size = 18)

p$NS + p$CTS + p$`2` + p$`5` + p$`10` +
  p$`15` + p$`20` + p$`25` + p$`30` +
  p$`35` + p$`40` + p$HK  +
  plot_layout(nrow = 3) & NoLegend() & coord_fixed() & NoAxis() 
rm(p)
```

### UMAP

For now, let's use all the PCs. In future iterations I expect this number to change.
Let's calculate the UMAP represenation

```{r}
dims = length(features)
set.seed(seed)
umap_data <- umap(d = pca_data$x[,1:dims], preserve.seed = TRUE)
```

Update the embeddings data frame

```{r}
data_embeddings <- cbind(regions, scaled_data, as.data.frame(pca_data$x))
data_embeddings <- cbind(data_embeddings, as.data.frame(umap_data$layout))
```

Visualize the CREs in the UMAP

```{r fig.width=4, fig.height=4}
p_size = 0.1
p <- PlotGroups(object = data_embeddings, group = "type", dim1 = "V1", dim2 = "V2",
                point_size = p_size, high_color = my_color,order = TRUE, text_size = 18)

p$NS + p$CTS + p$`2` + p$`5` + p$`10` +
  p$`15` + p$`20` + p$`25` + p$`30` +
  p$`35` + p$`40` + p$HK  +
  plot_layout(nrow = 3) & NoLegend() & coord_fixed() & NoAxis() 

# rm(p)
```

Features plots

```{r}
p_size = 0.1
tmp <- data_embeddings %>% mutate("Dim1" = V1, "Dim2" = V2) 
pf <- list()
for(n in colnames(scaled_data)){
  pf[[n]] <- tmp %>% myFeaturePlot(feature = n, order = TRUE, point_size = p_size, text_size = 14, high_color = my_color) + ggtitle(n)
}

p$NS + p$CTS + p$HK +
  pf[[1]] +  pf[[2]] + pf[[3]] +
  plot_layout(ncol = 3)  & 
  coord_fixed() & NoAxis() & theme(axis.line = element_line())
```

There is a cluster with low levels of the main Histone marks

### Clustering

Let's get those CREs with low levels of these three marks and high density of NS and CTS

```{r}
set.seed(777)
clusters <- ClusterLouvain(x = data_embeddings[,paste0("PC", 1:7)], resolution = 0.3, k = 30)
table(clusters)
data_embeddings <- data_embeddings %>% mutate(cluster = factor(clusters))
# rm(clusters)
```

Visualize the clusters in the UMAP plot

```{r fig.width=3}
data_embeddings %>% ggplot( mapping = aes(x = V1, y = V2, color = cluster)) + geom_point(size = p_size) + theme_bw() + coord_fixed() + LegendDotSize(size = 5)
```

Calculate the proportion of CREs by cell types across cluster

```{r fig.width=6, fig.height=2}
palette <- c("#88CCEE", "#CC6677", "#DDCC77", "#6699CC",   "#AA4499", 
                 "#888888", "#44AA99", "#999933", "#332288","#E69F00", "black", "blue", "gray")

p1 <- data_embeddings %>% group_by(cluster) %>% count(type) %>% ggplot(mapping = aes(x = cluster, y = n, fill = type)) + geom_bar(position = "fill", stat = "identity") + scale_fill_manual(values = palette)
p2 <- data_embeddings %>% group_by(type) %>% count(cluster) %>% ggplot(mapping = aes(x = type, y = n, fill = cluster)) + geom_bar(position = "fill", stat = "identity") + scale_fill_manual(values = palette)

p1 + p2 
```

The suspicious CREs are captured in cluster 5. 
More than half of the CREs in cluster 5 are Negative samples.  Let's remove this cluster and re-do the analysis.

### Second interation

Remove CREs from cluster 5 

```{r}
regions_filtered <- regions %>% filter(data_embeddings$cluster != 5)
rm(regions)
```

Re-scale the data -> calculate PCA 

I would like to save the scaled data separately for future analyses.

```{r}
scaled_data <- apply(X = regions_filtered[,features], MARGIN = 2, FUN = scale, center = TRUE, scale = TRUE)
colnames(scaled_data) <- paste0(colnames(scaled_data),"_scaled")
#PCA
set.seed(seed)
pca_data <- prcomp(x = scaled_data, center = FALSE, scale. = FALSE, rank. = length(features))
plot(y = pca_data$sdev, x = 1:length(pca_data$sdev))
```

For now, let's use all the PCs. In future iterations I expect this number to change.
Let's calculate the UMAP representation

```{r}
dims = length(features)
set.seed(seed)
umap_data <- umap(d = pca_data$x[,1:dims], preserve.seed = TRUE)
```

Update embeddings

```{r}
data_embeddings <- cbind(regions_filtered, scaled_data, as.data.frame(pca_data$x))
data_embeddings <- cbind(data_embeddings, as.data.frame(umap_data$layout))
```

Clustering

```{r}
set.seed(777)
clusters <- ClusterLouvain(x = data_embeddings[,paste0("PC", 1:7)], resolution = 0.3, k = 30)
table(clusters)
data_embeddings <- data_embeddings %>% mutate(cluster = factor(clusters))
# rm(clusters)
```

Visualize clusters

```{r fig.width=3}
data_embeddings %>% ggplot( mapping = aes(x = V1, y = V2, color = cluster)) + geom_point(size = p_size) + theme_bw() + coord_fixed() + LegendDotSize(size = 5)
```

Calculate the proportion of CREs by cell types across clusters

```{r fig.width=6, fig.height=2}
palette <- c("#88CCEE", "#CC6677", "#DDCC77", "#6699CC",   "#AA4499", 
                 "#888888", "#44AA99", "#999933", "#332288","#E69F00", "black", "blue", "gray")

p1 <- data_embeddings %>% group_by(cluster) %>% count(type) %>% ggplot(mapping = aes(x = cluster, y = n, fill = type)) + geom_bar(position = "fill", stat = "identity") + scale_fill_manual(values = palette)
p2 <- data_embeddings %>% group_by(type) %>% count(cluster) %>% ggplot(mapping = aes(x = type, y = n, fill = cluster)) + geom_bar(position = "fill", stat = "identity") + scale_fill_manual(values = palette)

p1 + p2 
```

Cluster 5 and 2 are suspicious. Let's analyze these clusters independently

### Sub-clustering analysis of cluster 5

Select CREs from cluster 5

```{r}
cl5 <- data_embeddings %>% filter(data_embeddings$cluster == 5)
table(cl5$n_celltypes)
```

Re-scale -> PCA

```{r}
scaled_data <- apply(X = cl5[,features], MARGIN = 2, FUN = scale, center = TRUE, scale = TRUE)
colnames(scaled_data) <- paste0(colnames(scaled_data),"_scaled")
#PCA
set.seed(seed)
pca_data <- prcomp(x = scaled_data, center = FALSE, scale. = FALSE, rank. = length(features))
plot(y = pca_data$sdev, x = 1:length(pca_data$sdev))
```

UMAP

```{r}
dims = length(features)
set.seed(seed)
umap_cl5 <- umap(d = pca_data$x[,1:dims], preserve.seed = TRUE)
```

Update embeddings of cluster 5

```{r}
cl5_embeddings <- cbind(cl5, scaled_data, as.data.frame(pca_data$x))
cl5_embeddings <- cbind(cl5_embeddings, as.data.frame(umap_cl5$layout))
```

Sub-clustering

```{r}
set.seed(777)
clusters_cl5 <- ClusterLouvain(x = cl5_embeddings[,paste0("PC", 1:7)], resolution = 0.3, k = 30)
table(clusters_cl5)
cl5_embeddings$cluster <- factor(clusters_cl5)
# rm(clusters)
```

Visualize clusters

```{r fig.width=3}
cl5_embeddings %>% ggplot( mapping = aes(x = V1, y = V2, color = cluster)) + geom_point(size = p_size) + theme_bw() + coord_fixed() + LegendDotSize(size = 5)
```

Proportion of CREs in clusters

```{r fig.width=6, fig.height=2}
palette <- c("#88CCEE", "#CC6677", "#DDCC77", "#6699CC",   "#AA4499", 
                 "#888888", "#44AA99", "#999933", "#332288","#E69F00", "black", "blue", "gray")

p1 <- cl5_embeddings %>% group_by(cluster) %>% count(type) %>% ggplot(mapping = aes(x = cluster, y = n, fill = type)) + geom_bar(position = "fill", stat = "identity") + scale_fill_manual(values = palette)
p2 <- cl5_embeddings %>% group_by(type) %>% count(cluster) %>% ggplot(mapping = aes(x = type, y = n, fill = cluster)) + geom_bar(position = "fill", stat = "identity") + scale_fill_manual(values = palette)

p1 + p2 
```

Select suspicious CREs

```{r}
cl5_rmv <- cl5_embeddings %>% filter(cluster != 1)
table(cl5_rmv$cluster)
```

let's filter suspicious CREs from the original data.

```{r}
data_embeddings <- data_embeddings %>% filter(!region_id %in% cl5_rmv$region_id)
rm(cl5, cl5_embeddings, cl5_rmv, clusters_cl5, umap_cl5)
```

```{r}
table(data_embeddings$cluster)
```

### Sub-clustering analysis of cluster 2

Let's do the same analysis for cluster 2 

```{r}
cl2 <- data_embeddings %>% filter(data_embeddings$cluster == 2)
table(cl2$n_celltypes)
```

remove previous PCA and umap

```{r}
rmv_idx <- which(colnames(cl2) %in% c(paste0("PC",1:7) ,"V1", "V2"))
cl2 <- cl2[,-rmv_idx]
```

Re-scale -> PCA

```{r}
scaled_data <- apply(X = cl2[,features], MARGIN = 2, FUN = scale, center = TRUE, scale = TRUE)
colnames(scaled_data) <- paste0(colnames(scaled_data),"_scaled")
#PCA
set.seed(seed)
pca_data <- prcomp(x = scaled_data, center = FALSE, scale. = FALSE, rank. = length(features))
plot(y = pca_data$sdev, x = 1:length(pca_data$sdev))
```

UMAP

```{r}
dims = length(features)
set.seed(seed)
umap_cl2 <- umap(d = pca_data$x[,1:dims], preserve.seed = TRUE)
```

Update embeddings

```{r}
cl2_embeddings <- cbind(cl2, as.data.frame(pca_data$x))
cl2_embeddings <- cbind(cl2_embeddings, as.data.frame(umap_cl2$layout))
```

Sub-clustering

```{r}
set.seed(777)
clusters_cl2 <- ClusterLouvain(x = cl2_embeddings[,paste0("PC", 1:7)], resolution = 0.3, k = 30)
table(clusters_cl2)
cl2_embeddings$cluster <- factor(clusters_cl2)
# rm(clusters)
```

Visualize clusters

```{r fig.width=3}
cl2_embeddings %>% ggplot( mapping = aes(x = V1, y = V2, color = cluster)) + geom_point(size = p_size) + theme_bw() + coord_fixed() + LegendDotSize(size = 5)
```

Proportion of clusters

```{r fig.width=6, fig.height=2}
palette <- c("#88CCEE", "#CC6677", "#DDCC77", "#6699CC",   "#AA4499", 
                 "#888888", "#44AA99", "#999933", "#332288","#E69F00", "black", "blue", "gray")

p1 <- cl2_embeddings %>% group_by(cluster) %>% count(type) %>% ggplot(mapping = aes(x = cluster, y = n, fill = type)) + geom_bar(position = "fill", stat = "identity") + scale_fill_manual(values = palette)
p2 <- cl2_embeddings %>% group_by(type) %>% count(cluster) %>% ggplot(mapping = aes(x = type, y = n, fill = cluster)) + geom_bar(position = "fill", stat = "identity") + scale_fill_manual(values = palette)

p1 + p2 
```

select suspicious regions

```{r}
cl2_rmv <- cl2_embeddings %>% filter(cluster == 1)
table(cl2_rmv$cluster)
```

remove suspicious regions from the original data.

```{r}
data_embeddings <- data_embeddings %>% filter(!region_id %in% cl2_rmv$region_id)
rm(cl2, cl2_embeddings, cl2_rmv, clusters_cl2, umap_cl2)
table(data_embeddings$type)
```

Short test
regions with zero signals for histones?
```{r}
sel_ann <- c("H3K27ac_n_cts", "H3K4me1_n_cts", "H3K4me3_n_cts")
sel_regions <- data_embeddings
for(ann in sel_ann){
  sel_regions <- sel_regions %>% filter(.data[[ann]] == 0)
}

sel_regions %>% MyHistogram(x = "n_celltypes", stat = "count") + scale_y_continuous(trans = "sqrt", breaks = c(0,100,1000,2000, 6000)) 

rm(ann, sel_ann, sel_regions)
```

Noup!!! this looks good. Let's get the final filtered regions on this loop 

```{r}
regions_filtered <- regions_filtered %>% filter(region_id %in% data_embeddings$region_id)
rm(data_embeddings)
```

## Save filtered regions

```{r}
saveRDS(object = regions_filtered, paste0(out_dir, "/u_regions_NS_filtered_2_", date, ".rds"))
```

