---
title: "Unsupervised analysis of annotated ABC + NS"
author: "Martin loza"
date: "2023/02/16"
---

In previous workflows I obtained diverse kind of annotations of the extended ABC + NS (e.g. regions +-500bps). The annotations are:
- Histone modifications
- Nucleotids content and CpG content
- Nearest TSS
- ABC annotations (number of cell types, number of target genes, etc.)
- Conservation score

I performed QC based on the percentage of Ns and the number of overlapping cell types for the histone marks.
I normalized the annotations. Each annotations was independently analysed and the type of normalization were chose accordingly.
I performed QC bases on PCA + clustering analysis

On this workflow I would like to perform the unsupervised analysis of the regions. The main goals is to annotate the regions and maybe finding something interesting.

```{r setup}
library(dplyr)
library(ggplot2)
library(patchwork)
library(umap)
library(bluster)
library(igraph)
library(scales)
library(viridis)
library(stringr)

## Global variables
date = "230216"
seed = 777
in_dir = "Data/annotated/"
out_dir = in_dir
my_color = "deepskyblue4"

## Global functions
source("Analysis/Functions.R")
source("Analysis/Functions_Visualizations.r")

#Local functions
```

## Load regions

Load annotations: histone, cpg, distance to nearest tss

```{r}
regions <- readRDS(file = paste0(in_dir, "/u_regions_NS_filtered_2_230215.rds"))

#short number of regions  
test <- table(regions$type)[c("CTS", "NS", "HK")] == c(48265,1397,10965)
if((!test[1]) | (!test[2])){
  cat("Number of regions do not match!!!!")
}else{
  cat("Number of NS, CTS, and HK regions OK")
}
rm(test)
```

## Set up

Selected features for unsupervised analysis

```{r}
features <- c("H3K27ac"  , "H3K4me1", "H3K4me3",
              "H3K27me3", "H3K36me3", "H3K9ac",
              "H3K9me3",
              "pct_GC_norm", "pct_AT_norm", "per_N_norm", "obsExp_CpG_norm",
              "dist_to_nearest_transcript_norm",
              "conservation_score_norm")

cat("Number of features: \n", length(features))
cat("\nNAs in features OK: \n", any(!is.na(regions[,features])))
```

Setup label for visualizations

```{r}
regions <- regions %>% mutate(type = "other") 
types <- c(1,2,5,10,15,20, 25, 30, 35, 40, 50, "NS")
for(t in types){
  #get indexed cells
  idx <- which(regions$n_celltypes == t)
  #set cts type
  if(t == "1"){
    regions$type[idx] <- "CTS"
  }else if(t == "50"){
    regions$type[idx] <- "HK"
  }else{
    regions$type[idx] <- t
  }
}
rm(idx)

regions$type <- factor(regions$type, levels = c("CTS",2,5,10,15,20, 25, 30, 35, 40, "HK", "NS", "other"))

table(regions$type)
rm(types, t)
```

## Iteration 3

### Scaled data

```{r}
scaled_data <- apply(X = regions[,features], MARGIN = 2, FUN = scale, center = TRUE, scale = TRUE)
colnames(scaled_data) <- paste0(colnames(scaled_data),"_scaled")

#short test
if(any(!is.numeric(scaled_data))){
  cat("Ohh no... there are non numeric elements in the data. ")
}else{
  cat("Data seems OK")
}
```

### PCA

Calculate PCA and visualize the variance captured by each PC

```{r}
set.seed(seed)
pca_data <- prcomp(x = scaled_data, center = FALSE, scale. = FALSE, rank. = length(features))
plot(y = pca_data$sdev, x = 1:length(pca_data$sdev))
```

Analyse the relation between PCs and features 

```{r}
pca_data$rotation
```

Update embeddings

```{r}
data_embeddings <- cbind(regions, scaled_data, as.data.frame(pca_data$x))
```

Visualize the CREs in the first 3 PCs

```{r fig.width=4}
p_size = 0.5
sel_regions <- data_embeddings %>%
  mutate(n_celltypes = factor(n_celltypes)) %>% arrange(n_celltypes)
p1 <- sel_regions %>% ggplot(mapping = aes(x = PC1, y = PC2)) + geom_point(size = p_size, colour = my_color) + theme_bw() + coord_fixed()
p2 <- sel_regions %>% ggplot(mapping = aes(x = PC1, y = PC3)) + geom_point(size = p_size, colour = my_color) + theme_bw() 
p3 <- sel_regions %>% ggplot(mapping = aes(x = PC2, y = PC3)) + geom_point(size = p_size, colour = my_color) + theme_bw() 

p1 + p2 + p3 + plot_layout(guides = "collect") 

rm(sel_regions,p1,p2,p3)
```

```{r fig.width=4, fig.height=4}
p_size = 0.1
p <- PlotGroups(object = data_embeddings, group = "type", dim1 = "PC1", dim2 = "PC2",
                point_size = p_size, high_color = my_color,order = TRUE, text_size = 18)

p$NS + p$CTS + p$`2` + p$`5` + p$`10` +
  p$`15` + p$`20` + p$`25` + p$`30` +
  p$`35` + p$`40` + p$HK  +
  plot_layout(nrow = 3) & NoLegend() & coord_fixed() & NoAxis() 
rm(p)
```

```{r fig.width=4, fig.height=4}
p_size = 0.1
p <- PlotGroups(object = data_embeddings, group = "type", dim1 = "PC1", dim2 = "PC3",
                point_size = p_size, high_color = my_color,order = TRUE, text_size = 18)

p$NS + p$CTS + p$`2` + p$`5` + p$`10` +
  p$`15` + p$`20` + p$`25` + p$`30` +
  p$`35` + p$`40` + p$HK  +
  plot_layout(nrow = 3) & NoLegend() & coord_fixed() & NoAxis() 
rm(p)
```

### UMAP

For now, let's use all the PCs. In future iterations I expect this number to change.
Let's calculate the UMAP represenation

```{r}
dims = length(features)
set.seed(seed)
umap_data <- umap(d = pca_data$x[,1:dims], preserve.seed = TRUE, method = 'umap-learn')
```

Update the embeddings data frame

```{r}
data_embeddings <- cbind(regions, scaled_data, as.data.frame(pca_data$x))
data_embeddings <- cbind(data_embeddings, as.data.frame(umap_data$layout))
```

Visualize the CREs in the UMAP

```{r fig.width=4, fig.height=3}
p_size = 0.5
sel_regions <- data_embeddings %>%
  # filter(n_celltypes == 1 | n_celltypes == 50 | n_celltypes == "NS") %>%
  mutate(n_celltypes = factor(n_celltypes)) %>% arrange(n_celltypes)
p1 <- sel_regions %>% ggplot(mapping = aes(x = PC1, y = PC2)) + geom_point(size = p_size, colour = my_color) + theme_bw() + coord_fixed()
p2 <- sel_regions %>% ggplot(mapping = aes(x = PC1, y = PC3)) + geom_point(size = p_size, colour = my_color) + theme_bw() 
p3 <- sel_regions %>% ggplot(mapping = aes(x = PC2, y = PC3)) + geom_point(size = p_size, colour = my_color) + theme_bw() 
p4 <- sel_regions %>% ggplot(mapping = aes(x = V1, y = V2)) + geom_point(size = p_size, colour = my_color) + theme_classic() + coord_fixed() +  NoAxis() + theme(axis.line = element_line()) 


((p1 + p2 + p3) / p4)  + 
  plot_layout(guides = "collect", nrow = 2, heights = c(1,1.5), widths = c(1,1,1)) 

rm(sel_regions,p1,p2,p3, p4)
```

Features plots

```{r}
p_size = 0.1
tmp <- data_embeddings %>% mutate("Dim1" = V1, "Dim2" = V2) 
pf <- list()
for(n in colnames(scaled_data)){
  pf[[n]] <- tmp %>% myFeaturePlot(feature = n, order = TRUE, point_size = p_size, text_size = 14, high_color = my_color) + ggtitle(n)
}
rm(tmp)
```

```{r fig.width=4, fig.height=4}
p_size = 0.1
p <- PlotGroups(object = data_embeddings, group = "type", dim1 = "V1", dim2 = "V2",
                point_size = p_size, high_color = my_color,order = TRUE, text_size = 18)

p$NS + p$CTS + p$`2` + p$`5` + p$`10` +
  p$`15` + p$`20` + p$`25` + p$`30` +
  p$`35` + p$`40` + p$HK  +
  plot_layout(nrow = 3) & NoLegend() & coord_fixed() & NoAxis(keep_lines = TRUE) 

# rm(p)
```

```{r fig.width=4, fig.height=4}
p$NS + p$CTS + p$HK +
  pf[[1]] +  pf[[2]] + pf[[3]] +
  pf[[4]] +  pf[[5]] + pf[[6]] + pf[[7]] +
  plot_layout(ncol = 3)  & 
  coord_fixed() & NoAxis(keep_lines = TRUE)
```

There are some islands which seems to have high density of NS and CTS regions. Let's use clustering analysis to capture them and analyse them.

### Clustering

```{r}
set.seed(777)
clusters <- ClusterLouvain(x = data_embeddings[,paste0("PC", 1:dims)],
                           resolution = 0.5, k = 30)
table(clusters)
data_embeddings <- data_embeddings %>% mutate(cluster = factor(clusters))
# rm(clusters)
```

Visualize the clusters in UMAP

```{r fig.width=3}
data_embeddings %>% ggplot( mapping = aes(x = V1, y = V2, color = cluster)) + geom_point(size = p_size) + theme_bw() + coord_fixed() + LegendDotSize(size = 5)
```

```{r fig.width=6, fig.height=2}
palette <- c("#88CCEE", "#CC6677", "#DDCC77", "#6699CC",   "#AA4499", 
                 "#888888", "#44AA99", "#999933", "#332288","#E69F00", "black", "blue", "gray", "deepskyblue4")

p1 <- data_embeddings %>% group_by(cluster) %>% count(type) %>% ggplot(mapping = aes(x = cluster, y = n, fill = type)) + geom_bar(position = "fill", stat = "identity") + scale_fill_manual(values = palette)
p2 <- data_embeddings %>% group_by(type) %>% count(cluster) %>% ggplot(mapping = aes(x = type, y = n, fill = cluster)) + geom_bar(position = "fill", stat = "identity") + scale_fill_manual(values = palette)

p1 + p2 
```

Let't perform independent analysis of suspicious clusters

### Analysis cluster 12

Let's check cluster 12

```{r}
clu_data <- regions %>% filter(data_embeddings$cluster == 12)
table(clu_data$type)
```

### Sub-clustering analysis


```{r}
#PCA
set.seed(seed)
pca_clu <- prcomp(x = clu_data[,features], center = TRUE, scale. = TRUE, rank. = length(features))
plot(y = pca_clu$sdev, x = 1:length(pca_clu$sdev))
#umap
dims = 10
set.seed(seed)
umap_clu <- umap(d = pca_clu$x[,1:dims], preserve.seed = TRUE, method = 'umap-learn')
```

Update the embeddings 

```{r}
clu_embeddings <- cbind(clu_data, as.data.frame(pca_clu$x))
clu_embeddings <- cbind(clu_embeddings, as.data.frame(umap_clu$layout))
```

Visuzualize CREs in umap

```{r fig.width=4, fig.height=4}
p_size = 0.1
p <- PlotGroups(object = clu_embeddings, group = "type", dim1 = "V1", dim2 = "V2",
                point_size = p_size, high_color = my_color,order = TRUE, text_size = 18)

p$NS + p$CTS + p$`2` + p$`5` + p$`10` +
  p$`15` + p$`20` + p$`25` + p$`30` +
  p$`35` + p$`40` + p$HK  +
  plot_layout(nrow = 3) & NoLegend() & coord_fixed() & NoAxis(keep_lines = TRUE) 

# rm(p)
```

Sub-clustering 

```{r}
set.seed(777)
clusters <- ClusterLouvain(x = clu_embeddings[,paste0("PC", 1:dims)], resolution = 0.1, k = 30)
table(clusters)
clu_embeddings <- clu_embeddings %>% mutate(cluster = factor(clusters))
# rm(clusters)
```

Visualize the sub-clusters
```{r fig.width=3}
clu_embeddings %>% ggplot( mapping = aes(x = V1, y = V2, color = cluster)) + geom_point(size = p_size) + theme_bw() + coord_fixed() + LegendDotSize(size = 5)
```

Visualize the proportion of CREs on each cluster

```{r fig.width=6, fig.height=2}
palette <- c("#88CCEE", "#CC6677", "#DDCC77", "#6699CC",   "#AA4499", 
                 "#888888", "#44AA99", "#999933", "#332288","#E69F00", "black", "blue", "gray")

p1 <- clu_embeddings %>% group_by(cluster) %>% count(type) %>% ggplot(mapping = aes(x = cluster, y = n, fill = type)) + geom_bar(position = "fill", stat = "identity") + scale_fill_manual(values = palette)
p2 <- clu_embeddings %>% group_by(type) %>% count(cluster) %>% ggplot(mapping = aes(x = type, y = n, fill = cluster)) + geom_bar(position = "fill", stat = "identity") + scale_fill_manual(values = palette)

p1 + p2 
```

Select and remove suspicious clusters
Let's remove the cluster 1 of cluster 12. It contains mostly NS, CTS, and 2 regions.

```{r}
clu_rmv <- clu_embeddings %>% filter(cluster == 1)
table(clu_rmv$type)
```

Filter suspicious regions from embeddings and regions data frames

```{r}
data_embeddings <- data_embeddings %>% filter(!region_id %in% clu_rmv$region_id)
rm(clu_data, clu_embeddings, clu_rmv, pca_clu, umap_clu)
```

```{r}
regions <- regions %>% filter(region_id %in% data_embeddings$region_id)
```

Now let's analyse cluster 2

### Analysis cluster 2

```{r}
clu_data <- regions %>% filter(data_embeddings$cluster == 2)
table(clu_data$type)
```

### Re analysis

PCA -> UMAP

```{r}
#PCA
set.seed(seed)
pca_clu <- prcomp(x = clu_data[,features], center = TRUE, scale. = TRUE, rank. = length(features))
plot(y = pca_clu$sdev, x = 1:length(pca_clu$sdev))
dims = 10
set.seed(seed)
umap_clu <- umap(d = pca_clu$x[,1:dims], preserve.seed = TRUE, method = 'umap-learn')
```

Update the embeddings data frame

```{r}
clu_embeddings <- cbind(clu_data, as.data.frame(pca_clu$x))
clu_embeddings <- cbind(clu_embeddings, as.data.frame(umap_clu$layout))
```

Visuzualize the CREs in the UMAP

```{r fig.width=4, fig.height=4}
p_size = 0.1
p <- PlotGroups(object = clu_embeddings, group = "type", dim1 = "V1", dim2 = "V2",
                point_size = p_size, high_color = my_color,order = TRUE, text_size = 18)

p$NS + p$CTS + p$`2` + p$`5` + p$`10` +
  p$`15` + p$`20` + p$`25` + p$`30` +
  p$`35` + p$`40` + p$HK  +
  plot_layout(nrow = 3) & NoLegend() & coord_fixed() & NoAxis(keep_lines = TRUE) 

# rm(p)
```

Sub-clustering

```{r}
set.seed(777)
clusters <- ClusterLouvain(x = clu_embeddings[,paste0("PC", 1:dims)], resolution = 0.1, k = 30)
table(clusters)
clu_embeddings <- clu_embeddings %>% mutate(cluster = factor(clusters))
# rm(clusters)
```

Visualize clusters in the UMAP 

```{r fig.width=3}
clu_embeddings %>% ggplot( mapping = aes(x = V1, y = V2, color = cluster)) + geom_point(size = p_size) + theme_bw() + coord_fixed() + LegendDotSize(size = 5)
```

```{r fig.width=6, fig.height=2}
palette <- c("#88CCEE", "#CC6677", "#DDCC77", "#6699CC",   "#AA4499", 
                 "#888888", "#44AA99", "#999933", "#332288","#E69F00", "black", "blue", "gray")

p1 <- clu_embeddings %>% group_by(cluster) %>% count(type) %>% ggplot(mapping = aes(x = cluster, y = n, fill = type)) + geom_bar(position = "fill", stat = "identity") + scale_fill_manual(values = palette)
p2 <- clu_embeddings %>% group_by(type) %>% count(cluster) %>% ggplot(mapping = aes(x = type, y = n, fill = cluster)) + geom_bar(position = "fill", stat = "identity") + scale_fill_manual(values = palette)

p1 + p2 
```

Filter suspicious clusters which contains mostly NS, CTS, and 2 regions.

```{r}
clu_rmv <- clu_embeddings %>% filter(cluster == 1)
table(clu_rmv$type)
```

Update embeddings and regions 

```{r}
data_embeddings <- data_embeddings %>% filter(!region_id %in% clu_rmv$region_id)
rm(clu_data, clu_embeddings, clu_rmv, pca_clu, umap_clu)
regions <- regions %>% filter(region_id %in% data_embeddings$region_id)
```

Now let's analyse cluster 3

### Analysis cluster 3

```{r}
clu_data <- regions %>% filter(data_embeddings$cluster == 3)
table(clu_data$type)
```

### Re analysis

PCA -> UMAP

```{r}
#PCA
set.seed(seed)
pca_clu <- prcomp(x = clu_data[,features], center = TRUE, scale. = TRUE, rank. = length(features))
plot(y = pca_clu$sdev, x = 1:length(pca_clu$sdev))
dims = 11
set.seed(seed)
umap_clu <- umap(d = pca_clu$x[,1:dims], preserve.seed = TRUE, method = 'umap-learn')
```

Update embeddings

```{r}
clu_embeddings <- cbind(clu_data, as.data.frame(pca_clu$x))
clu_embeddings <- cbind(clu_embeddings, as.data.frame(umap_clu$layout))
```

Visualize embeddings

```{r fig.width=4, fig.height=4}
p_size = 0.1
p <- PlotGroups(object = clu_embeddings, group = "type", dim1 = "V1", dim2 = "V2",
                point_size = p_size, high_color = my_color,order = TRUE, text_size = 18)

p$NS + p$CTS + p$`2` + p$`5` + p$`10` +
  p$`15` + p$`20` + p$`25` + p$`30` +
  p$`35` + p$`40` + p$HK  +
  plot_layout(nrow = 3) & NoLegend() & coord_fixed() & NoAxis(keep_lines = TRUE) 

# rm(p)
```

Sub-clustering

```{r}
set.seed(777)
clusters <- ClusterLouvain(x = clu_embeddings[,paste0("PC", 1:dims)], resolution = 0.2, k = 30)
table(clusters)
clu_embeddings <- clu_embeddings %>% mutate(cluster = factor(clusters))
# rm(clusters)
```

Visualize clusters in UMAP

```{r fig.width=3}
clu_embeddings %>% ggplot( mapping = aes(x = V1, y = V2, color = cluster)) + geom_point(size = p_size) + theme_bw() + coord_fixed() + LegendDotSize(size = 5)
```

```{r fig.width=6, fig.height=2}
palette <- c("#88CCEE", "#CC6677", "#DDCC77", "#6699CC",   "#AA4499", 
                 "#888888", "#44AA99", "#999933", "#332288","#E69F00", "black", "blue", "gray")

p1 <- clu_embeddings %>% group_by(cluster) %>% count(type) %>% ggplot(mapping = aes(x = cluster, y = n, fill = type)) + geom_bar(position = "fill", stat = "identity") + scale_fill_manual(values = palette)
p2 <- clu_embeddings %>% group_by(type) %>% count(cluster) %>% ggplot(mapping = aes(x = type, y = n, fill = cluster)) + geom_bar(position = "fill", stat = "identity") + scale_fill_manual(values = palette)

p1 + p2 
```

Filter suspicious clusters which contains mostly NS, CTS, and 2 regions.

```{r}
clu_rmv <- clu_embeddings %>% filter(cluster == 2)
table(clu_rmv$type)
```

Update embeddings and regions

```{r}
data_embeddings <- data_embeddings %>% filter(!region_id %in% clu_rmv$region_id)
rm(clu_data, clu_embeddings, clu_rmv, pca_clu, umap_clu)
regions <- regions %>% filter(region_id %in% data_embeddings$region_id)
```

Now let's analyse cluster 6

### Analysis cluster 6

```{r}
clu_data <- regions %>% filter(data_embeddings$cluster == 6 )
table(clu_data$type)
```
### Re analysis

PCA -> UMAP

```{r}
#PCA
set.seed(seed)
pca_clu <- prcomp(x = clu_data[,features], center = TRUE, scale. = TRUE, rank. = length(features))
plot(y = pca_clu$sdev, x = 1:length(pca_clu$sdev))
dims = 8
set.seed(seed)
umap_clu <- umap(d = pca_clu$x[,1:dims], preserve.seed = TRUE, method = 'umap-learn')
```

Update embeddings data frame

```{r}
clu_embeddings <- cbind(clu_data, as.data.frame(pca_clu$x))
clu_embeddings <- cbind(clu_embeddings, as.data.frame(umap_clu$layout))
```

Visualize embeddings

```{r fig.width=4, fig.height=4}
p_size = 0.5
p <- PlotGroups(object = clu_embeddings, group = "type", dim1 = "V1", dim2 = "V2",
                point_size = p_size, high_color = my_color,order = TRUE, text_size = 18)

p$NS + p$CTS + p$`2` + p$`5` + p$`10` +
  p$`15` + p$`20` + p$`25` + p$`30` +
  p$`35` + p$`40` + p$HK  +
  plot_layout(nrow = 3) & NoLegend() & coord_fixed() & NoAxis(keep_lines = TRUE) 

# rm(p)
```

Sub-clustering

```{r}
set.seed(777)
clusters <- ClusterLouvain(x = clu_embeddings[,paste0("PC", 1:dims)], resolution = 0.3, k = 30)
table(clusters)
clu_embeddings <- clu_embeddings %>% mutate(cluster = factor(clusters))
# rm(clusters)
```

Visualize the clusters in the UMAP plot

```{r fig.width=3}
clu_embeddings %>% ggplot( mapping = aes(x = V1, y = V2, color = cluster)) + geom_point(size = p_size) + theme_bw() + coord_fixed() + LegendDotSize(size = 5)
```

Calculate the proportion of CREs by cell types across cluster

```{r fig.width=6, fig.height=2}
palette <- c("#88CCEE", "#CC6677", "#DDCC77", "#6699CC",   "#AA4499", 
                 "#888888", "#44AA99", "#999933", "#332288","#E69F00", "black", "blue", "gray")

p1 <- clu_embeddings %>% group_by(cluster) %>% count(type) %>% ggplot(mapping = aes(x = cluster, y = n, fill = type)) + geom_bar(position = "fill", stat = "identity") + scale_fill_manual(values = palette)
p2 <- clu_embeddings %>% group_by(type) %>% count(cluster) %>% ggplot(mapping = aes(x = type, y = n, fill = cluster)) + geom_bar(position = "fill", stat = "identity") + scale_fill_manual(values = palette)

p1 + p2 
```

Remove suspicious clusters which contains mostly NS, CTS, and 2 regions.

```{r}
clu_rmv <- clu_embeddings %>% filter(cluster == 3)
table(clu_rmv$type)
```

Filter suspicious CREs

```{r}
data_embeddings <- data_embeddings %>% filter(!region_id %in% clu_rmv$region_id)
rm(clu_data, clu_embeddings, clu_rmv, pca_clu, umap_clu)
regions <- regions %>% filter(region_id %in% data_embeddings$region_id)
```

## Iteration 4

### Scaled data

```{r}
scaled_data <- apply(X = regions[,features], MARGIN = 2, FUN = scale, center = TRUE, scale = TRUE)
colnames(scaled_data) <- paste0(colnames(scaled_data),"_scaled")

#short test
if(any(!is.numeric(scaled_data))){
  cat("Ohh no... there are non numeric elements in the data. ")
}else{
  cat("Data seems OK")
}
```

### PCA

Let's calcualte the PCA and confirm the variance captured.
```{r}
set.seed(seed)
pca_data <- prcomp(x = scaled_data, center = FALSE, scale. = FALSE, rank. = length(features))
plot(y = pca_data$sdev, x = 1:length(pca_data$sdev))
```

Analyse the relation between PCs and features

```{r}
pca_data$rotation
```

Update embedding data frame

```{r}
data_embeddings <- cbind(regions, scaled_data, as.data.frame(pca_data$x))
```

Visualize the CREs in the first 3 PCs

```{r fig.width=4}
p_size = 0.5
sel_regions <- data_embeddings %>%
  mutate(n_celltypes = factor(n_celltypes)) %>% arrange(n_celltypes)
p1 <- sel_regions %>% ggplot(mapping = aes(x = PC1, y = PC2)) + geom_point(size = p_size, colour = my_color) + theme_bw() + coord_fixed()
p2 <- sel_regions %>% ggplot(mapping = aes(x = PC1, y = PC3)) + geom_point(size = p_size, colour = my_color) + theme_bw() 
p3 <- sel_regions %>% ggplot(mapping = aes(x = PC2, y = PC3)) + geom_point(size = p_size, colour = my_color) + theme_bw() 

p1 + p2 + p3 + plot_layout(guides = "collect") 

rm(sel_regions,p1,p2,p3)
```

```{r fig.width=4, fig.height=4}
p_size = 0.1
p <- PlotGroups(object = data_embeddings, group = "type", dim1 = "PC1", dim2 = "PC2",
                point_size = p_size, high_color = my_color,order = TRUE, text_size = 18)

p$NS + p$CTS + p$`2` + p$`5` + p$`10` +
  p$`15` + p$`20` + p$`25` + p$`30` +
  p$`35` + p$`40` + p$HK  +
  plot_layout(nrow = 3) & NoLegend() & coord_fixed() & NoAxis() 
rm(p)
```

```{r fig.width=4, fig.height=4}
p_size = 0.1
p <- PlotGroups(object = data_embeddings, group = "type", dim1 = "PC1", dim2 = "PC3",
                point_size = p_size, high_color = my_color,order = TRUE, text_size = 18)

p$NS + p$CTS + p$`2` + p$`5` + p$`10` +
  p$`15` + p$`20` + p$`25` + p$`30` +
  p$`35` + p$`40` + p$HK  +
  plot_layout(nrow = 3) & NoLegend() & coord_fixed() & NoAxis() 
rm(p)
```

### UMAP

For now, let's use all the PCs. In future iterations I expect this number to change.
Let's calculate the UMAP represenation

```{r}
dims = length(features)
set.seed(seed)
umap_data <- umap(d = pca_data$x[,1:dims], preserve.seed = TRUE, method = 'umap-learn')
```

Update the embeddings data frame

```{r}
data_embeddings <- cbind(regions, scaled_data, as.data.frame(pca_data$x))
data_embeddings <- cbind(data_embeddings, as.data.frame(umap_data$layout))
```

Visualize the CREs in the UMAP

```{r fig.width=4, fig.height=3}
p_size = 0.5
sel_regions <- data_embeddings %>%
  # filter(n_celltypes == 1 | n_celltypes == 50 | n_celltypes == "NS") %>%
  mutate(n_celltypes = factor(n_celltypes)) %>% arrange(n_celltypes)
p1 <- sel_regions %>% ggplot(mapping = aes(x = PC1, y = PC2)) + geom_point(size = p_size, colour = my_color) + theme_bw() + coord_fixed()
p2 <- sel_regions %>% ggplot(mapping = aes(x = PC1, y = PC3)) + geom_point(size = p_size, colour = my_color) + theme_bw() 
p3 <- sel_regions %>% ggplot(mapping = aes(x = PC2, y = PC3)) + geom_point(size = p_size, colour = my_color) + theme_bw() 
p4 <- sel_regions %>% ggplot(mapping = aes(x = V1, y = V2)) + geom_point(size = p_size, colour = my_color) + theme_classic() + coord_fixed() +  NoAxis() + theme(axis.line = element_line()) 


((p1 + p2 + p3) / p4)  + 
  plot_layout(guides = "collect", nrow = 2, heights = c(1,1.5), widths = c(1,1,1)) 

rm(sel_regions,p1,p2,p3, p4)
```

Features plots

```{r}
p_size = 0.1
tmp <- data_embeddings %>% mutate("Dim1" = V1, "Dim2" = V2) 
pf <- list()
for(n in colnames(scaled_data)){
  pf[[n]] <- tmp %>% myFeaturePlot(feature = n, order = TRUE, point_size = p_size, text_size = 14, high_color = my_color) + ggtitle(n)
}
rm(tmp)
```

```{r fig.width=4, fig.height=4}
p_size = 0.1
p <- PlotGroups(object = data_embeddings, group = "type", dim1 = "V1", dim2 = "V2",
                point_size = p_size, high_color = my_color,order = TRUE, text_size = 18)

p$NS + p$CTS + p$`2` + p$`5` + p$`10` +
  p$`15` + p$`20` + p$`25` + p$`30` +
  p$`35` + p$`40` + p$HK  +
  plot_layout(nrow = 3) & NoLegend() & coord_fixed() & NoAxis(keep_lines = TRUE) 

# rm(p)
```

```{r fig.width=4, fig.height=4}
p$NS + p$CTS + p$HK +
  pf[[1]] +  pf[[2]] + pf[[3]] +
  pf[[4]] +  pf[[5]] + pf[[6]] + pf[[7]] +
  plot_layout(ncol = 3)  & 
  coord_fixed() & NoAxis(keep_lines = TRUE)
```

There are some islands which seems to have high density of NS and CTS regions. Let's use clustering analysis to capture them and analyse them.

### Clustering

```{r}
set.seed(777)
clusters <- ClusterLouvain(x = data_embeddings[,paste0("PC", 1:dims)],
                           resolution = 0.5, k = 30)
table(clusters)
data_embeddings <- data_embeddings %>% mutate(cluster = factor(clusters))
# rm(clusters)
```

Visualize the clusters in the UMAP plot

```{r fig.width=3}
data_embeddings %>% ggplot( mapping = aes(x = V1, y = V2, color = cluster)) + geom_point(size = p_size) + theme_bw() + coord_fixed() + LegendDotSize(size = 5)
```

Calculate the proportion of CREs by cell types across cluster

```{r fig.width=6, fig.height=2}
palette <- c("#88CCEE", "#CC6677", "#DDCC77", "#6699CC",   "#AA4499", 
                 "#888888", "#44AA99", "#999933", "#332288","#E69F00", "black", "blue", "gray", "deepskyblue4")

p1 <- data_embeddings %>% group_by(cluster) %>% count(type) %>% ggplot(mapping = aes(x = cluster, y = n, fill = type)) + geom_bar(position = "fill", stat = "identity") + scale_fill_manual(values = palette)
p2 <- data_embeddings %>% group_by(type) %>% count(cluster) %>% ggplot(mapping = aes(x = type, y = n, fill = cluster)) + geom_bar(position = "fill", stat = "identity") + scale_fill_manual(values = palette)

p1 + p2 
```

Let's investigate suspicious clusters

### Analysis cluster 7

```{r}
clu_data <- regions %>% filter(data_embeddings$cluster == 7)
table(clu_data$type)
```

PCA -> UMAP

```{r}
#PCA
set.seed(seed)
pca_clu <- prcomp(x = clu_data[,features], center = TRUE, scale. = TRUE, rank. = length(features))
plot(y = pca_clu$sdev, x = 1:length(pca_clu$sdev))
dims = 13
set.seed(seed)
umap_clu <- umap(d = pca_clu$x[,1:dims], preserve.seed = TRUE, method = 'umap-learn')
```

Update embeddings

```{r}
clu_embeddings <- cbind(clu_data, as.data.frame(pca_clu$x))
clu_embeddings <- cbind(clu_embeddings, as.data.frame(umap_clu$layout))
```

Visualize the CREs in the UMAP

```{r fig.width=4, fig.height=4}
p_size = 0.1
p <- PlotGroups(object = clu_embeddings, group = "type", dim1 = "V1", dim2 = "V2",
                point_size = p_size, high_color = my_color,order = TRUE, text_size = 18)

p$NS + p$CTS + p$`2` + p$`5` + p$`10` +
  p$`15` + p$`20` + p$`25` + p$`30` +
  p$`35` + p$`40` + p$HK  +
  plot_layout(nrow = 3) & NoLegend() & coord_fixed() & NoAxis(keep_lines = TRUE) 

# rm(p)
```

Sub-clustering

```{r}
set.seed(777)
clusters <- ClusterLouvain(x = clu_embeddings[,paste0("PC", 1:dims)], resolution = 0.3, k = 30)
table(clusters)
clu_embeddings <- clu_embeddings %>% mutate(cluster = factor(clusters))
# rm(clusters)
```

Visualize the clusters in the UMAP plot

```{r fig.width=3}
clu_embeddings %>% ggplot( mapping = aes(x = V1, y = V2, color = cluster)) + geom_point(size = p_size) + theme_bw() + coord_fixed() + LegendDotSize(size = 5)
```

Calculate the proportion of CREs by cell types across cluster

```{r fig.width=6, fig.height=2}
palette <- c("#88CCEE", "#CC6677", "#DDCC77", "#6699CC",   "#AA4499", 
                 "#888888", "#44AA99", "#999933", "#332288","#E69F00", "black", "blue", "gray")

p1 <- clu_embeddings %>% group_by(cluster) %>% count(type) %>% ggplot(mapping = aes(x = cluster, y = n, fill = type)) + geom_bar(position = "fill", stat = "identity") + scale_fill_manual(values = palette)
p2 <- clu_embeddings %>% group_by(type) %>% count(cluster) %>% ggplot(mapping = aes(x = type, y = n, fill = cluster)) + geom_bar(position = "fill", stat = "identity") + scale_fill_manual(values = palette)

p1 + p2 
```

Let's remove suspicious regions in clusters that contains mostly NS, CTS, and 2 regions.

```{r}
clu_rmv <- clu_embeddings %>% filter(cluster == 1)
table(clu_rmv$type)
```

Filter suspicious regions from embeddings and regions

```{r}
data_embeddings <- data_embeddings %>% filter(!region_id %in% clu_rmv$region_id)
rm(clu_data, clu_embeddings, clu_rmv, pca_clu, umap_clu)
regions <- regions %>% filter(region_id %in% data_embeddings$region_id)
```

## Save data

```{r}
saveRDS(object = regions, paste0(out_dir, "/u_regions_NS_filtered_3_", date, ".rds"))
```




























































































































































































































































































