---
title: "Re-normalization of annotated ABC + NS"
author: "Martin loza"
date: "2023/03/04"
---

In previous workflow I filter out regions using unsupervised analyses and NS. On this workflow I would like to re-normalize the data for further annotation analysis.

```{r setup}
library(dplyr)
library(ggplot2)
library(patchwork)

## Global variables
date = "230304"
seed = 777
in_dir = "Data/annotated/"
out_dir = in_dir

## Global functions
source("Analysis/Functions.R")
source("Analysis/Functions_Visualizations.r")

#Local functions

```

## Load regions

Load annotated regions 

```{r}
regions <- readRDS(file = paste0(in_dir, "/u_regions_NS_filtered_3_230216.rds"))

#short number of regions  
test <- table(regions$type)[c("CTS", "NS", "HK")] == c(31700,431,10965)
if((!test[1]) | (!test[2])){
  cat("Number of regions do not match!!!!")
}else{
  cat("Number of NS, CTS, and HK regions OK")
}
rm(test)
```

```{r}
regions %>% MyHistogram(x = "n_celltypes", stat = "count") + theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

## Normalization

We analized each features independently and how different normalization affect their distribution. For more info check the Github repository containing the whole set of test/results/etc.

The next normalization/scaling will be performed

- Histone marks: nothing
- Distance to nearest TSS: abs + log1p

```{r}
#normalize the distance to nearest TSS: abs + log1p
sel_ann <- c("dist_to_nearest_transcript")
for(ann in sel_ann){
  label = paste0(ann, "_norm")
  regions[[label]] <- log1p(abs(regions[[ann]]))
}
rm(sel_ann, ann)
```

## Save normalized regions

```{r}
saveRDS(object = regions, paste0(out_dir, "/u_regions_NS_filtered_normalized_", date, ".rds"))
```













