---
title: "Setup medium embeddings"
author: "Martin loza"
date: "2023/03/04"
---

In previous workflows I performed QC based on the percentage of Ns and the number of overlapping cell types for the histone markk and re-normalized the annotations.

On this workflow I would like to scale the features for further analyses.

```{r setup}
library(dplyr)
library(ggplot2)
library(patchwork)
library(umap)
library(bluster)
library(igraph)
library(scales)
library(viridis)
library(stringr)

## Global variables
date = "230304"
seed = 777
in_dir = "Data/"
out_dir = in_dir
my_color = "deepskyblue4"

## Global functions
source("Analysis/Functions.R")
source("Analysis/Functions_Visualizations.r")

#Local functions
```

## Load regions

Load annotations: histone signals, and distance to nearest tss

```{r}
regions <- readRDS(file = paste0(in_dir, "/annotated/u_regions_NS_filtered_normalized_230304.rds"))

#short number of regions  
test <- table(regions$type)[c("CTS", "NS", "HK")] == c(31700,431,10965)
if((!test[1]) | (!test[2])){
  cat("Number of regions do not match!!!!")
}else{
  cat("Number of NS, CTS, and HK regions OK")
}
rm(test)
```

```{r}
regions %>% MyHistogram(x = "n_celltypes", stat = "count") + theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

## Set up

Verify the type labels

```{r}
table(regions$type)
```

Verify the scaled features. Because I remove many regions, it would be nice to re-scale the data just in case

```{r}
sel_columns <- c("H3K27ac", "H3K4me1", "H3K4me3",
                 "dist_to_nearest_transcript_norm")
#short tests
cat("Number of features: \n", length(sel_columns))
cat("\nAll features in the regions' data frame OK: \n")
sum(sel_columns %in% colnames(regions)) == length(sel_columns)
```

remove previous scaled data if existing

```{r}
idx <- which(str_detect(colnames(regions), pattern = "scaled"))
if(length(idx) != 0){
  regions <- regions[,-idx]
}
colnames(regions)
rm(idx)
```

## Scaling

Let's scale the data and re-assign them in the data frame

```{r}
#scale the features
scaled_data <- scale(x = regions[, sel_columns], center = TRUE, scale = TRUE)
#rename the columns
colnames(scaled_data) <- paste0(colnames(scaled_data), "_scaled")
#short test. All the columns of scaled data in regions
t <- sum(colnames(scaled_data) %in% colnames(regions)) == ncol(scaled_data)
#add scaled data to regions
regions <- cbind(regions,scaled_data)
colnames(regions)
rm(t, scaled_data, sel_columns)
```

## Save data
```{r}
saveRDS(object = regions, file = paste0(out_dir,"/annotated/u_regions_NS_filtered_scaled_",date,".rds"))
```

Check the scaled features
```{r fig.width=5}
sel_regions <- regions
sel_ann <- c("H3K27ac_scaled", "H3K4me1_scaled", "H3K4me3_scaled")
fill = NULL
plots <- list()

for(ann in sel_ann){
  label <- paste0("Normalized ",ann)
  plots[[ann]] <- sel_regions %>%  MyHistogram(x = ann) + ggtitle(label)
}
(plots[[1]] + plots[[2]] + plots[[3]] + plots[[4]] + plots[[5]] + plots[[6]] + plots[[7]] & scale_y_continuous(trans = "sqrt")) + plot_layout(ncol = 3, guides = "collect")
rm(label, ann,  fill, plots, sel_regions, sel_ann)
```

```{r fig.width=4}
sel_regions <- regions
sel_ann <- paste0(c("dist_to_nearest_transcript"), "_norm_scaled")
fill = NULL
plots <- list()

for(ann in sel_ann){
  label <- paste0("Normalized ",ann)
  plots[[ann]] <- sel_regions %>%  MyHistogram(x = ann) + ggtitle(label)
}
(plots[[1]] + plots[[2]] & scale_y_continuous(trans = "sqrt")) + plot_layout(ncol = 2, guides = "collect")
rm(label, ann,  fill, plots, sel_regions, sel_ann)
```
