---
title: "Setup embeddings"
author: "Martin loza"
date: "2023/03/04"
---

In previous workflows I performed QC based on the percentage of Ns and the number of overlapping cell types for the histone marks.
I normalized the annotations and scaled them after QC using NS and unsupervised analysis.

On this workflow I would like to setup the embeddings to use in further studies. Calculating embeddings take a long time so it is a good idea to set them up so I don't have to calculated them every time I make a new analysis.

```{r setup}
library(dplyr)
library(ggplot2)
library(patchwork)
library(umap)
library(bluster)
library(igraph)
library(scales)
library(viridis)
library(stringr)

## Global variables
date = "230304"
seed = 777
in_dir = "Data/"
out_dir = in_dir
my_color = "deepskyblue4"

## Global functions
source("Analysis/Functions.R")
source("Analysis/Functions_Visualizations.r")

#Local functions
```

## Load regions

Load annotations: histone, cpg, distance to nearest tss

```{r}
regions <- readRDS(file = paste0(in_dir, "/annotated/u_regions_NS_filtered_scaled_230304.rds"))

#short number of regions  
test <- table(regions$type)[c("CTS", "NS", "HK")] == c(31700,431,10965)
if((!test[1]) | (!test[2])){
  cat("Number of regions do not match!!!!")
}else{
  cat("Number of NS, CTS, and HK regions OK")
}
rm(test)
```

```{r}
regions %>% MyHistogram(x = "n_celltypes", stat = "count") + theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

## Set up

Verify the type labels

```{r}
table(regions$type)
```

Let's remove any columns related to PCA or UMAP
```{r}
#remove columns related to PC. Umap columns are also related to UMAP
idx <- which(str_detect(string = colnames(regions), pattern = "PC"))
if(length(idx) != 0){
  regions <- regions[,-idx]
}
idx <- which(str_detect(string = colnames(regions), pattern = "UMAP"))
if(length(idx) != 0){
  regions <- regions[,-idx]
}
colnames(regions)
rm(idx)
```

### Embeddings

Select features

```{r}
features <- c("H3K27ac_scaled", "H3K4me1_scaled", "H3K4me3_scaled",
              "dist_to_nearest_transcript_norm_scaled")
#short tests
cat("Number of features: \n", length(features))
cat("\nAll features in the data frame OK:\n", sum(features %in% colnames(regions)) == length(features) ) 
cat("\nNAs in features OK: \n", any(!is.na(regions[,features])))
```

Get PCA coordinates

```{r}
#get PCA
set.seed(seed)
pca_data <- prcomp(x = regions[,features], center = FALSE, scale. = FALSE, rank. = length(features))
#elbow plot
plot(y = pca_data$sdev, x = 1:length(pca_data$sdev))
#rotation
pca_data$rotation
```

Select the coordinates and rename the columns: "low"(resolution) 

```{r}
pca_embeddings <- as.data.frame(pca_data$x)
colnames(pca_embeddings)
rm(pca_data)
```

Get UMAP coordinates

```{r}
dims = length(features)
set.seed(seed)
umap_data <- umap(d = pca_embeddings[,1:dims], preserve.seed = TRUE, method = 'umap-learn')
#setup umap embeddings
umap_embeddings <- as.data.frame(umap_data$layout)
colnames(umap_embeddings) <- c("UMAP_1", "UMAP_2")
embeddings <- cbind(pca_embeddings, umap_embeddings)
rm(pca_embeddings, umap_embeddings, umap_data, dims)
```

Setup embeddings before saving

```{r}
rownames(embeddings) <- regions$region_id
```

Save embeddings

```{r}
saveRDS(object = embeddings, file = paste0(in_dir, "/embeddings/low_embeddings_", date, ".rds"))
```













