---
title: "Annotation of low embeddings"
author: "Martin loza"
date: "2023/03/05"
---


In previous workflows I performed QC based on the number of overlapping cell types for the histone marks. I normalized the annotations. Each annotations was independently analysed and the type of normalization were chose accordingly.
I performed QC bases on PCA + clustering analysis and I obtained the low dimensionality embeddings. 

On this workflow I would like to perform the unsupervised analysis of the low embeddings. The main goals are to annotate the regions and maybe finding something interesting.

```{r setup}
library(dplyr)
library(ggplot2)
library(patchwork)
library(umap)
library(bluster)
library(igraph)
library(scales)
library(viridis)
library(stringr)

## Global variables
date = "230305"
seed = 777
in_dir = "Data/"
out_dir = in_dir
my_color <- "deepskyblue4"
palette <- c("#88CCEE", "#CC6677", "#DDCC77", "#6699CC",   "#AA4499", 
                 "#888888", "#44AA99", "#999933", "#332288","#E69F00", "black", "blue", "gray", "deepskyblue4")

## Global functions
source("Analysis/Functions.R")
source("Analysis/Functions_Visualizations.r")

#Local functions

RelabelClusters <- function(.data = NULL, clusters = NULL, old_labels = NULL, new_labels = NULL, new_clusters = NULL){
  #check that clusters are in .data
  if(!clusters %in% colnames(.data)){
    stop(call. = TRUE, "ERROR: The clusters labels were not found.")
  }
  
  #check that the size of old and new labels match
  if(length(old_labels) != length(new_labels)){
    stop(call. = TRUE, "ERROR: The number of labels doesn't match.")
  }
  
  #check that the clusters are a factor
  if(!"factor" %in% is(.data[[clusters]])){
    stop(call. = TRUE, "ERROR: The clusters should be a factor.")
  }
  
   #factor's levels are stored as characters. For comparisons let's transform the old/new labels to character
  old_labels <- as.character(old_labels)
  new_labels <- as.character(new_labels)
  
  #if the new_clusters is NULL, we assign the new labels in the previous factor
  if(is.null(new_clusters)){
    new_clusters <- clusters
  }
  
  #get the old clusters for comparison
  old_clusters <- as.character(.data[[clusters]])
  #init the new clusters
  ncl <- as.character(.data[[clusters]])
  
  for(i in seq_along(old_labels)){
      
    #get the index of the current element
    idx <- which(old_clusters == old_labels[i])
    #assign the new clusters
    ncl[idx] <- new_labels[i]
  }
    
  #if the new and old cluster has the same name, we need to remove the previous clusters
  if(new_clusters == clusters){
    .data <- .data[,-which(colnames(.data) == clusters)]
  }
  
  #new colnames
  new_colnames <- c(colnames(.data), new_clusters)
  #assign the new clusters
  .data <- cbind(.data,factor(x = ncl))
  #assisng colnames
  colnames(.data) <- new_colnames

  return(.data)
}

RemoveColumn <- function(.data = NULL, column = NULL){
  #ut. Columns in colnames
  if(sum(!column %in% colnames(.data)) == length(column)){
    stop(call. = TRUE, "ERROR: Columns not found in the data.")
  }
  idx <- which(colnames(.data) %in% column)
  .data <- .data[,-idx]
  return(.data)
}

```

## Load regions

Load annotations: histone, cpg, distance to nearest tss

```{r}
regions <- readRDS(file = paste0(in_dir, "/annotated/u_regions_NS_filtered_scaled_230304.rds"))
embeddings <- readRDS(file = paste0(in_dir, "/embeddings/low_embeddings_230304.rds"))

#short number of regions  
test <- table(regions$type)[c("CTS", "NS", "HK")] == c(31700,431,10965)
if((!test[1]) | (!test[2])){
  cat("Number of regions do not match!!!!")
}else{
  cat("Number of NS, CTS, and HK regions OK")
}
rm(test)
```

```{r}
regions %>% MyHistogram(x = "n_celltypes", stat = "count") + theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

## Set up

Merge the regions and their embeddings 

```{r}
#short test
ut <- identical(rownames(embeddings), regions$region_id)
cat("Regions_id are identical? \n", ut)
if(ut){
  regions <- cbind(regions, embeddings)
}
rm(ut, embeddings)
```

In this analysis I will use only 4 main features

```{r}
#original features
features <- c("H3K27ac"  , "H3K4me1", "H3K4me3",
              "dist_to_nearest_transcript_norm")
#scaled features
features <- paste0(features, "_scaled")

cat("Number of features: \n", length(features))
cat("\nAll features in the data frame OK:\n", sum(features %in% colnames(regions)) == length(features) ) 
cat("\nNAs in features OK: \n", any(!is.na(regions[,features])))
```

## Unsupervised analysis

## Main distribution of the data 

```{r fig.width=4, fig.height=4}
p_size = 0.1
#regions by number of cell types
umap_columns <- c("UMAP_1",  "UMAP_2")
p <- PlotGroups(object = regions, group = "type",
                dim1 = umap_columns[1],
                dim2 = umap_columns[2],
                point_size = p_size, high_color = my_color,order = TRUE, text_size = 18)

#regions by features
tmp <- regions %>% mutate("Dim1" = UMAP_1, "Dim2" = UMAP_2) 
pf <- list()
for(n in features){
  pf[[n]] <- tmp %>% myFeaturePlot(feature = n, order = TRUE, point_size = p_size, text_size = 14, high_color = my_color) + NoAxis() + scale_colour_viridis_c(option = "magma") + ggtitle(n)
}
rm(tmp, n, p_size)
```

```{r fig.width=7}
plot <- p$NS + p$CTS + p$HK +
  pf[[1]] +  pf[[2]] + pf[[3]] +
  pf[[4]] +
  plot_layout(ncol = 3)  & 
  coord_fixed() & NoAxis()

plot
rm(pf,p, plot)
```

Let's try to classify enhancer/promoter-like regions.

## Clustering high level

```{r}
#we have 4 dimensions
dims = 4
set.seed(777)
clusters <- ClusterLouvain(x = regions[,paste0("PC", 1:dims)],
                           resolution = 0.2, k = 30)
table(clusters)
regions <- regions %>% mutate(clusters_louvain_0.2 = factor(clusters))
# rm(clusters,dims)
```

Visualize the clusters in the UMAP

```{r}
p_size = 0.1
regions %>% ggplot( mapping = aes(x = UMAP_1, y = UMAP_2, color = clusters_louvain_0.2)) + geom_point(size = p_size) + theme_bw() + coord_fixed() + LegendDotSize(size = 2)
rm(p_size)
``` 

By comparing with the features plots we can roughly classify the clusters as :

- Cluster 1,2,4: Main_positive (three main features are positive in this cluster)
- Cluster 3: H3K27ac negative
- Cluster 5: H3K4me3 negative
- Cluster 6: H3K4me1 negative

Let's relabel the clusters
```{r}
regions<- regions %>% RelabelClusters(clusters = "clusters_louvain_0.2", new_clusters = "ann_v0",
                                      old_labels = c(1,2,4,3,5,6),
                                      new_labels = c("positive", "positive", "positive",
                                                     "H3K27ac_negative", "H3K4me3_negative","H3K4me1_negative"))
table(regions$ann_v0)
```

Let's check the re-labelled clusters

```{r}
p_size = 0.1
regions %>% ggplot( mapping = aes(x = UMAP_1, y = UMAP_2, color = ann_v0)) + geom_point(size = p_size) + theme_bw() + coord_fixed() + LegendDotSize(size = 2)
rm(p_size)
```

## Annotation of positive regions

```{r}
sel_regions <- subset(x = regions, subset = ann_v0 == "positive")
table(sel_regions$ann_v0)
```

Let's rescale and recalculate the PCA

```{r}
dims = 4
tmp_features <- c("H3K27ac","H3K4me1","H3K4me3","dist_to_nearest_transcript_norm")
#scaling
sel_scaled <- scale(x = sel_regions[,tmp_features], center = TRUE, scale = TRUE)
colnames(sel_scaled) <- paste0(tmp_features, "_scaled")
#PCA
set.seed(seed)
sel_pca <- prcomp(x = sel_scaled, center = FALSE, scale. = FALSE, rank. = dims)
plot(y = sel_pca$sdev, x = 1:length(sel_pca$sdev))
sel_pca$rotation
```

Let's recalculate the UMAP

```{r}
set.seed(seed)
sel_umap <- umap(d = sel_pca$x[,1:dims], preserve.seed = TRUE, method = 'umap-learn')
colnames(sel_umap$layout) <- c("UMAP_1", "UMAP_2")
rm(dims)
```

Remove previous columns in selected regions 

```{r}
sel_regions <- sel_regions %>% RemoveColumn(column = c(paste0(tmp_features, "_scaled"), #scaled features 
                                                       paste0("PC", 1:dims), #pca
                                                       c("UMAP_1", "UMAP_2"))) #umap
# colnames(sel_regions)
```

Let's assign the re-analized data 

```{r}
sel_regions <- cbind(sel_regions, sel_scaled, sel_pca$x, sel_umap$layout)
# colnames(sel_regions)
```

Re-clustering

```{r}
set.seed(777)
clusters <- ClusterLouvain(x = sel_regions[,paste0("PC", 1:dims)],
                           resolution = 0.3, k = 30)
table(clusters)
sel_regions <- sel_regions %>% mutate(tmp_clusters = factor(clusters))
rm(clusters)
```

Plot cluster's 

```{r fig.width=4}
p_size = 0.1
#regions by number of cell types
umap_columns <- c("UMAP_1",  "UMAP_2")
p <- PlotGroups(object = sel_regions, group = "tmp_clusters",
                dim1 = umap_columns[1],
                dim2 = umap_columns[2],
                point_size = p_size, high_color = my_color,order = TRUE, text_size = 18)
for(i in seq_along(p)){
  if(i == 1){
    tmp_p <- p[[i]]
  }else{
    tmp_p <- tmp_p + p[[i]]
  }
}

(tmp_p & theme_bw() & coord_fixed() &NoLegend()) + plot_layout(ncol = 3)

rm(tmp, n, p_size)
```

Plot clusters

```{r fig.width= 6}
p_size = 0.1
sel_regions %>% ggplot( mapping = aes(x = UMAP_1, y = UMAP_2, color = tmp_clusters)) + geom_point(size = p_size) + theme_bw() + coord_fixed() + LegendDotSize(size = 2)
rm(p_size)
```

Plot features' violin plots

```{r fig.width= 5}
features <- c("H3K27ac", "H3K4me3","H3K4me1", "dist_to_nearest_transcript_norm")
p <- list()
for(f in features){
   p[[f]] <- sel_regions %>%
     MyViolinPlot(x = "tmp_clusters", y = f, fill_violin = "tmp_clusters", 
                  fill = "white",box_width = 0.2) + ggtitle(f)
}

(p[[1]] + p[[2]] + p[[3]] + p[[4]] + plot_layout(guides = "collect", ncol = 2)) &theme(text = element_text(size = 16))
rm(p, f)
```

Print summaries
```{r}
f <- features[4]
cat(f,"\n")
for(c in unique(sel_regions$tmp_clusters)){
  cat("\nCluster ", c, " : \n")
  print(summary( expm1((sel_regions %>% filter(tmp_clusters == c))[[f]])))
}

```

Plot groups and feature plots

```{r fig.width=4, fig.height=4}
p_size = 0.1
#regions by number of cell types
umap_columns <- c("UMAP_1",  "UMAP_2")
p <- PlotGroups(object = sel_regions, group = "type",
                dim1 = umap_columns[1],
                dim2 = umap_columns[2],
                point_size = p_size, high_color = my_color,order = TRUE, text_size = 18)

#regions by features
tmp <- sel_regions %>% mutate("Dim1" = UMAP_1, "Dim2" = UMAP_2) 
pf <- list()
for(n in paste0(features, "_scaled")){
  pf[[n]] <- tmp %>% myFeaturePlot(feature = n, order = TRUE, point_size = p_size, text_size = 14, high_color = my_color) + NoAxis() + scale_colour_viridis_c(option = "magma") + ggtitle(n)
}
rm(tmp,n, p_size)
```

```{r fig.width=7}
plot <- p$NS + p$CTS + p$HK +
  pf[[1]] +  pf[[2]] + pf[[3]] +
  pf[[4]] +
  plot_layout(ncol = 3)  & 
  coord_fixed() & NoAxis()
plot
rm(pf,p, plot)
```

I think an initial classification would be:

-C1: Enhancers_medDist: H3K27ac+, H3K4me1++, dist+  (median = 4429)
-C2: Other: main histones negative, dist+ (median = 4618)
-C3: Enhancer_longDist: H3K4me3--, H3K4me1+, dist++ (median = 16,706)
-C4: Promoter_core: H3K27ac+, H3K4me3+, dist-- (median = 43)
-C5: Promoter: H3K27ac++, H3K4me3++, dist- (median = 944)

Let's re-label the clusters
```{r}
sel_regions <- sel_regions %>% RelabelClusters(clusters = "tmp_clusters",
                                               old_labels = c(1,2,3,4,5), 
                                               new_labels = c("Enhancer_medDist",
                                                              "Other",
                                                              "Enhancer_longDist",
                                                              "Promoter_core",
                                                              "Promoter"))
```

Let's check the plots again

Plot cluster's 

```{r fig.width=4}
p_size = 0.1
#regions by number of cell types
umap_columns <- c("UMAP_1",  "UMAP_2")
p <- PlotGroups(object = sel_regions, group = "tmp_clusters",
                dim1 = umap_columns[1],
                dim2 = umap_columns[2],
                point_size = p_size, high_color = my_color,order = TRUE, text_size = 18)
for(i in seq_along(p)){
  if(i == 1){
    tmp_p <- p[[i]]
  }else{
    tmp_p <- tmp_p + p[[i]]
  }
}

(tmp_p & theme_bw() & coord_fixed() &NoLegend()) + plot_layout(ncol = 3)

rm(tmp, n, p_size)
```

Plot features' violin plots

```{r fig.width= 5}
features <- c("H3K27ac", "H3K4me3","H3K4me1", "dist_to_nearest_transcript_norm")
p <- list()
for(f in features){
   p[[f]] <- sel_regions %>%
     MyViolinPlot(x = "tmp_clusters", y = f, fill_violin = "tmp_clusters", 
                  fill = "white",box_width = 0.2) + ggtitle(f)
}

(p[[1]] + p[[2]] + p[[3]] + p[[4]] + plot_layout (ncol = 2)) & NoLegend() &
  theme(text = element_text(size = 16), axis.text.x = element_text(angle = 50, hjust = 1))
rm(p, f)
```
 
```{r fig.width=6, fig.height=2}
p1 <- sel_regions %>% group_by(tmp_clusters) %>% count(type) %>% ggplot(mapping = aes(x = tmp_clusters, y = n, fill = type)) + geom_bar(position = "fill", stat = "identity") + scale_fill_manual(values = palette)
p2 <- sel_regions %>% group_by(type) %>% count(tmp_clusters) %>% ggplot(mapping = aes(x = type, y = n, fill = tmp_clusters)) + geom_bar(position = "fill", stat = "identity") + scale_fill_manual(values = palette)

p1 + p2 
```
 
They look OK.
 
Let's transfer these labels to the original regions.

```{r}
#let's create a temporal label for now
regions$tmp_label <- as.character(regions$ann_v0)
#in a simple way, let's check every region and transfer the new labels
for(i in seq_len(nrow(sel_regions))){
  #current region_id
  id <- sel_regions[i,"region_id"]
  #get the id's index in the larger data frame 
  idx <- which(regions$region_id == id)
  #assign the label
  regions[idx,"tmp_label"] <- as.character(sel_regions[i,"tmp_clusters"])
}

rm(id, idx)
```

```{r}
table(regions$ann_v0)
table(regions$tmp_label)
```

Let's replace the ann_v0 annotations

```{r}
regions$ann_v0 <- factor(regions$tmp_label)
regions <- regions %>% RemoveColumn(column = "tmp_label")
table(regions$ann_v0)
```

## Annotation v0

Plot features' violin plots

```{r fig.width= 5}
features <- c("H3K27ac", "H3K4me3","H3K4me1", "dist_to_nearest_transcript_norm")
p <- list()
for(f in features){
   p[[f]] <- regions %>%
     MyViolinPlot(x = "ann_v0", y = f, fill_violin = "ann_v0", 
                  fill = "white",box_width = 0.2) + ggtitle(f)
}

(p[[1]] + p[[2]] + p[[3]] + p[[4]] + plot_layout (ncol = 2)) & NoLegend() &
  theme(text = element_text(size = 16), axis.text.x = element_text(angle = 50, hjust = 1))
rm(p, f)
```

Plot annotations

```{r fig.width=6}
p_size = 0.1
#regions by number of cell types
umap_columns <- c("UMAP_1",  "UMAP_2")
p <- PlotGroups(object = regions , group = "ann_v0",
                dim1 = umap_columns[1],
                dim2 = umap_columns[2],
                point_size = p_size, high_color = my_color,order = TRUE, text_size = 18)
for(i in seq_along(p)){
  if(i == 1){
    tmp_p <- p[[i]]
  }else{
    tmp_p <- tmp_p + p[[i]]
  }
}

(tmp_p & theme_bw() & coord_fixed() &NoLegend() & NoAxis(keep_lines = TRUE)) + plot_layout(ncol = 4)

rm(tmp, n, p_size)
```

```{r fig.width=6, fig.height=2}
p1 <- regions %>% group_by(ann_v0) %>% count(type) %>% ggplot(mapping = aes(x = ann_v0, y = n, fill = type)) + geom_bar(position = "fill", stat = "identity") + scale_fill_manual(values = palette)
p2 <- regions %>% group_by(type) %>% count(ann_v0) %>% ggplot(mapping = aes(x = type, y = n, fill = ann_v0)) + geom_bar(position = "fill", stat = "identity") + scale_fill_manual(values = palette)

p1 + p2 
```

save first version of annotations

```{r}
saveRDS(object = regions,
        file = "Data/annotated_filtered/regions_annotated_v0_230306.rds")
```






