---
title: "Merge annotations"
author: "Martin loza"
date: "2023/02/15"
---

In previous notebooks I annotated the histones modification and nucleotide content of ABC + NS regions. In this workflow I would like to merge the annotations, including the one from the ABC merging procedure. 

```{r setup}
library(dplyr)
library(GenomicRanges)
library(stringr)
library(ggplot2)

## Global variables
date = "230215"
seed = 777
in_dir = "Data/annotated/"
out_dir = in_dir

## Global functions
source("Analysis/Functions.R")
source("Analysis/Functions_Visualizations.r")

#Local functions
```

## Load data 

Load ABC + NS regions

```{r}
regions <- readRDS(file = paste0(in_dir, "/merged_u_regions_NS_230210.rds"))

#short number of regions  
test <- table(regions$type) == c(225140,10000)
if((!test[1]) | (!test[2])){
  cat("Number of regions do not match!!!!")
}else{
  cat("Number of NS + ABC regions OK")
}
rm(test)
```

Load histones' annotations

```{r}
histone_ann <- readRDS(file = paste0(out_dir, "/histones_annotated_230215.rds"))
#short test
cat("Same number of rows OK: \n")
nrow(histone_ann) == nrow(regions)
cat("Same region_ids OK: \n")
sum(histone_ann$region_id %in% regions$region_id) == nrow(histone_ann)
```

Load nearest tss' annotations

```{r}
tss_ann <- readRDS(file = paste0(out_dir, "/nearest_tss_annotated_230210.rds"))
#short test
cat("Same number of rows OK: \n")
nrow(tss_ann) == nrow(regions)
cat("Same region_ids OK: \n")
sum(tss_ann$region_id %in% regions$region_id) == nrow(tss_ann)
```

Load ABC regions 

```{r}
abc_ann <- readRDS(file = "Data/merged/merged_nearest_gene_221130.rds")
abc_ann <- GetUniqueRegions(abc_ann)
#short test
cat("Same number of rows OK: \n")
nrow(abc_ann)+10000 == nrow(regions)
cat("Same region_ids OK: \n")
sum(abc_ann$region_id %in% regions$region_id) == nrow(abc_ann)
```
Final test

```{r}
#same number of elements
nrow(tss_ann) == nrow(histone_ann)
nrow(tss_ann) == (nrow(abc_ann) + 10000) #we have 10k NS
```

## Setup

The only annotation that has a different number of elements is the abc_ann, which only included the annotation of the abc elements, e.g. doesn't include the NS

Therefore, I would select the interesting annotations from the ABC enhancers.
I will set the extra rows in the NS as "NS"

```{r}
# colnames(abc_ann)
#I would like to conserve: regions_id, n_celltypes, test_n_celltypes, raw_CTSE
sel_cols <- c("region_id", "n_celltypes", "test_n_celltypes", "raw_CTSE", "n_genes")
abc_ann <- abc_ann %>% select(all_of(sel_cols))
rm(sel_cols)
colnames(abc_ann)
```

Now let's create the annotation for the remaining NS

```{r}
ns_ann <- as.data.frame(matrix(data = "NS", nrow = 10000, ncol = ncol(abc_ann)))
colnames(ns_ann) <- colnames(abc_ann)
#get NS region_id
ids <- regions %>% filter(type == "NS") %>% pull(region_id)
ns_ann <- ns_ann %>% mutate(region_id = ids)
#merge the abc_ann + ns_ann
abc_ann <- rbind(abc_ann, ns_ann)
#short test
cat("Same regions ids:\n")
sum(abc_ann$region_id %in% regions$region_id) == nrow(abc_ann)
rm(ids,ns_ann)
```

Now, let's use the region_ids as anchors to transfer the columns
For this, I create a function to ease the transfer

```{r}
#transfer the histone annotations
cols_idx <- which(!colnames(histone_ann) %in% colnames(regions))
sel_cols <- colnames(histone_ann)[cols_idx]
regions <- TransferColumns(reference = regions, query = histone_ann, 
                           anchors = "region_id", cols_to_tranfer = sel_cols)
rm(cols_idx, sel_cols)
```

```{r}
#transfer the tss annotations
cols_idx <- which(!colnames(tss_ann) %in% colnames(regions))
sel_cols <- colnames(tss_ann)[cols_idx]
regions <- TransferColumns(reference = regions, query = tss_ann, 
                           anchors = "region_id", cols_to_tranfer = sel_cols)
rm(cols_idx, sel_cols)
```

```{r}
#transfer the abc annotations
cols_idx <- which(!colnames(abc_ann) %in% colnames(regions))
sel_cols <- colnames(abc_ann)[cols_idx]
regions <- TransferColumns(reference = regions, query = abc_ann, 
                           anchors = "region_id", cols_to_tranfer = sel_cols)
rm(cols_idx, sel_cols)
```

Final columns

```{r}
colnames(regions)
```

Yay!! let's save the annotated data

```{r}
saveRDS(object = regions, file = paste0(out_dir,"u_regions_NS_annotated_",date,".rds" ))
```














































































