---
title: "Define universe"
author: "Martin loza"
date: "2022/11/14"
---

In a previous notebook I matched the annotations from ENSEMBL genes and ABC target genes. Even though I couldn't match around 1000 target genes, the majority of the ABC target were properly matched.

In this notebook I aim to define a gene and enhancers universe to use in further analyses.

NOTE. 
- The regions were already filtered by ABC.score >= 0.15.
- Regions and target's tss were lifted up to hg38. 
- Target genes were updated

These universes will be used in downstream analyses of the ABC enhancers.

```{r setup}
library(here)
library(dplyr)

## Global variables
date = "221121"
seed = 777
in_dir <- "Data/"
out_dir = in_dir
```

## Load data

Load the regions

```{r}
edata <- readRDS(file = paste0(in_dir,"/raw_enhancers/merged/03_1_hg38_genes_updated_221117.rds"))
```

Load the ENSEMBL transcripts and genes

```{r}
genes <- readRDS(file ="Data/Commonly_used/hg38/genes_ENSEMBL_hg38.rds")
transcripts <- readRDS(file ="Data/Commonly_used/hg38/transcripts_ENSEMBL_hg38.rds")
```

Add tss to transcripts and genes
```{r}
genes <- genes %>% mutate(tss = if_else(condition = strand == "+", true = start, false = end))
transcripts <- transcripts %>% mutate(tss = if_else(condition = strand == "+", true = start, false = end))
```

Do we have duplicated gene names, e.g. same gene with more than one version.

```{r}
sum(duplicated(genes$gene_name))
```

Yes, there are 1200 duplicated elements.

Let's transform the version into integer values

```{r}
genes <- genes %>% mutate(gene_version = as.integer(gene_version))
```

Let's get the duplicated genes

```{r}
dup_gene_names <- genes %>% filter(duplicated(gene_name)) %>% pull(gene_name) %>% unique()
head(dup_gene_names)
```

Let's split the genes into unique and duplicated 

```{r}
unique_genes <- genes %>% filter(!gene_name %in% dup_gene_names)
dup_genes <- genes %>% filter(gene_name %in% dup_gene_names)
```

Let's choose the highest version of duplicated genes

Choose the highest version
```{r}
# for each duplicated gene
for(g in dup_gene_names){
  #get the duplicated rows
  dup_rows <- dup_genes %>% filter(gene_name == g)
  #select the highest version
  sel_row <- dup_rows %>% filter(gene_version == max(gene_version))
  #if more than one row, e.g. more than one row matching with the highest version of the gene
  if(nrow(sel_row) > 1){
    #select the first row
    sel_row <- sel_row[1,]
  }

  #if filter_results does not exist
  if(!exists("filter_results")){
    #init filter_results as the sel_row
    filter_results <- sel_row
  }else{
    #update the results datasets
    filter_results <- rbind(filter_results, sel_row)
  }
}#end
rm(sel_row, dup_rows, g, dup_gene_names, dup_genes)
```

update the genes dataset

```{r}
# merged unique genes and the genes resulting after filtering duplicated genes
genes <- rbind(unique_genes, filter_results)
rm(unique_genes, filter_results)
```

Test again

```{r}
sum(duplicated(genes$gene_name))
```

Short test

All the gene names are in the transcripts' gene names

```{r}
sum(genes$gene_name %in% transcripts$gene_name) == nrow(genes)
```

Yes, no problem.

## Define universe of enhancers

```{r}
#get abc genes
abc_genes <- unique(edata$TargetGene)
#get ensembl_genes
ensemble_genes <- unique(genes$gene_name)
cat("Number of genes in ABC data:\n")
length(abc_genes)
cat("Number of genes in ENSEMBL data:\n")
length(ensemble_genes)
cat("Number of matching genes:\n")
sum(abc_genes %in% ensemble_genes)
```
Let's get the matching gene_names

```{r}
matching_genes <- intersect(ensemble_genes, abc_genes)
```

## Define the regions universe 

```{r}
original_abc_number <- nrow(edata)
cat("Original number of regions in the ABC data:\n")
original_abc_number
edata <- edata %>% filter(TargetGene %in% matching_genes)
cat("Number of regions after filtering:\n")
nrow(edata)
cat("Pecentage retained:\n")
paste0(100*round(nrow(edata)/original_abc_number,digits = 5), "%")
```

We lost around 3% of the abc data. I think this is acceptable

## Define the regions universe

```{r}
#get pcg in ensemble
ensemble_pcg <- genes %>% filter(gene_biotype == "protein_coding") %>% pull(gene_name) %>% unique()
cat("Number of PCG in ENSEMBLE data:\n")
length(ensemble_pcg)
cat("Number of ENSEMBL PCG in the genes-universe:\n")
sum(ensemble_pcg %in% matching_genes)
cat("Percentage of PCG:\n")
paste0(100*round(sum(ensemble_pcg %in% abc_genes)/length(ensemble_pcg), digits = 5), "%")
```

We will loose around 6% of the PCG annotated in ENSEMBL... This is kind of OK.

Let's get this project's genes and transcripts universe

```{r}
genes <- genes %>% filter(gene_name %in% matching_genes)
transcripts <- transcripts %>% filter(gene_name %in% matching_genes)
```

Universe distribution

```{r}
genes %>% filter(!duplicated(gene_name)) %>% pull(gene_biotype) %>%  table() %>% sort(decreasing = TRUE) 
```

Final check
```{r}
sum(unique(edata$TargetGene)%in% genes$gene_name) == length(unique(edata$TargetGene))
sum(genes$gene_name %in% edata$TargetGene) == nrow(genes)
sum(unique(transcripts$gene_name) %in% edata$TargetGene) == length(unique(transcripts$gene_name))
```
# Save data

```{r}
saveRDS(object = edata, file = paste0(out_dir, "/raw_enhancers/merged/03_2_hg38_regions_universe_", date, ".rds"))
saveRDS(object = genes, file = paste0(out_dir,"/hg38/hg38_genes_universe_", date,".rds"))
saveRDS(object = transcripts, file = paste0(out_dir,"/hg38/hg38_transcripts_universe_", date,".rds"))
```


