---
title: "Analysis of cell types"
author: "Martin Loza"
date: "2022/11/14"
---

In this notebook I aim to analyse the cell-types/tissues included in the ABC's database to filter cancer cells or cells with some treatment. 

```{r setup}
library(here)
library(dplyr)
library(stringr)

## Global variables
date = "221114"
out_dir = "/Data/"
```

## Load data

Download the ABC cell types information. This data is supplied in the supplementary table 2 from the publication: https://www.nature.com/articles/s41586-021-03446-x

Download the table and save it as ABC_celltypes_info.tsv

```{r}
ct_md_raw <- read.table(file = here("Data/Nasser/ABC_celltypes_info.tsv"), sep = "\t", header = TRUE)
head(ct_md_raw$ABC_celltype_id)
```

The cell-type information is in the "ABC_celltype_id" column.

## Filter cells 

### Find treated cells 

We would like to filter cells with some treatment.
Then, we will search for some keywords as "treated", "hour", and "hr" .
Also I found some treated cells that can be identified with the string "ATAC"

```{r}
ct_names <- ct_md_raw$ABC_celltype_id
```

```{r}
#Get the cell's index to remove
idx <- Reduce(f = union, x = list(which(str_detect(string = ct_names, pattern = "treated")),
                                  which(str_detect(string = ct_names, pattern = "hour")),
                                  which(str_detect(string = ct_names, pattern = "hr-")),
                                  which(str_detect(string = ct_names, pattern = "ATAC"))))

```

### Find cancer cells 

```{r}
# Manually selected cancer cells 
cancer_cells <- c("K562-Roadmap", "HepG2-Roadmap", "MCF-7-ENCODE",
                  "Panc1-ENCODE", "HCT116-ENCODE", "Jurkat-Engreitz",
                  "SK-N-SH-ENCODE", "GM12878-Roadmap", "A673-ENCODE", "BJAB-Engreitz", "HAP1", "HT29", "HeLa-S3-Roadmap",
                  "Karpas-422-ENCODE", "LNCAP", "LoVo", "MCF10A-Ji2017", "MDA-MB-231", "MM.1S-ENCODE", "NCCIT", 
                  "OCI-LY7-ENCODE", "PC-9-ENCODE", "THP-1_macrophage-VanBortle2017", "THP-1_monocyte-VanBortle2017",
                  "THP1-Engreitz", "U937-Engreitz")
# Get the index of cancer cells to remove
idx <- union(idx, which(ct_names %in%  cancer_cells))
```

```{r}
#check the cells to filter
ct_names[idx]
```

## Filter cells 

```{r}
sel_ct_md <- ct_md_raw %>% filter(ABC_celltype_id %in% ct_names[-idx])
sel_ct_md$ABC_celltype_id
```

## Save selected cells meta-data

```{r}
saveRDS(object = sel_ct_md, 
        file = paste0(out_dir, "selected_ABC_ct_", date, ".rds"))
```

