---
title: "Filter ABC cell types"
author: "Martin Loza"
data: "2022/11/14"
---

In a previous notebook named "04_1_Filter_ABC_celltypes" I filtered cancer and treated cell types/tissues from the original ABC database.

In this notebook I aim to filter cancer and treated cells from the enhancers data frame.

NOTE. 
- The regions were already filtered by ABC.score >= 0.15.
- Regions and target's tss were lifted up to hg38. 
- Target genes were annotated
- Regions targeting genes not found in ENSEMBLE annotations were filtered.

## Init

```{r setup}
library(dplyr)
library(here)

## Global variables
date = "221114"
#Set the files' directory (Data's folder by cell type)
in_dir <- "Data/"
```

## Load data

Load filtered cell types meta-data
```{r}
ct_md <- readRDS(file = paste0(in_dir, "/selected_ABC_ct_221114.rds"))
head(ct_md)
```

Load the enhancers data 

```{r}
edata <- readRDS(file = paste0(in_dir,"/raw_enhancers/merged/03_raw_enhancers_hg38_gene_annotated_221114.rds"))
```

## Filter cancer and treated cells's regions

```{r}
#save initial number of regions
initial_n_regions <- nrow(edata)
#filter cancer and treated cell types
edata <- edata %>% filter(CellType %in% ct_md$ABC_celltype_id)
```

```{r}
cat("Initial number of enhances:\n")
initial_n_regions
cat("Number of enhancers after filtering cancer and treated cell types:\n")
nrow(edata)
cat("Percentage of filtered regions from the original data sets:\n")
cat(100-100*round(nrow(edata)/initial_n_regions, digits = 6)," %" )
```
We filtered around 45% of the cell types... this is expected as we filtered 60 cell types from 131 original ones.

## Save filtered regions

```{r}
saveRDS(object = edata, 
        file = paste0(in_dir,"/raw_enhancers/merged/04_raw_enhancers_hg38_gene_annotated_cts_filtered_",date,".rds"))
```



