---
title: "Select regions from 50 cts"
author: "Martin loza"
date: "2022/11/14"
---

In a previous analysis I filtered cancer and treated cell types/tissues from the original ABC enhancers.

In this notebook I aim to randomly select 50 cell types to use in further analyses.

NOTE. 
- The regions were already filtered by ABC.score >= 0.15.
- Regions and target's tss were lifted up to hg38. 
- Target genes were annotated
- Regions targeting genes not found in ENSEMBLE annotations were filtered.
- Regions targeting cancer or treated cell types were filtered. 

```{r setup}
library(dplyr)
library(GenomicRanges)
library(here)
library(bedr)
library(ggplot2)

## Global variables
date = "221114"
seed = 777
in_dir <- "Data/"
out_dir = in_dir
```

## Load data

Load enhancers data 

```{r}
enhancers <- readRDS(file = paste0(in_dir, "/raw_enhancers/merged/04_raw_enhancers_hg38_gene_annotated_cts_filtered_221114.rds"))
```

## Select cell types

Randomly select 50 cell types

```{r}
#set number of cell types
n = 50
#get unique cell types labels
celltypes <- unique(enhancers$CellType)
#set seed for reproducibility and randomly select the idx of cell types
set.seed(seed)
sel_celltypes <- sample(x = celltypes, size = n, replace = FALSE)
#get current selected enhancers
sel_enhancers <- enhancers %>% filter(CellType %in% sel_celltypes)
cat("Selected cell types: \n")
unique(enhancers$CellType)
```

## Short analysis with protein coding genes

Load ENSEMBL annotated genes

```{r}
ensembl_genes <- readRDS(file = "/Volumes/MARTIN_LOZA/Projects/Enhancer_Grammar/Data/Commonly_used/hg38/hg38_genes_all_annotated_221113.rds")
```

```{r}
#get genes targeted in the selected dataset
sel_data_genes <- sel_enhancers %>% filter(!duplicated(TargetGene)) %>% select(TargetGene,gene_biotype)
cat("Biotype of selected regions:\n")
sel_data_genes %>% pull(gene_biotype) %>% table() %>% sort(decreasing = TRUE)
```

The selected regions target 17592 protein coding genes

```{r}
cat("Number of protein coding genes in the ENSEMBL gene dataset:\n")
ensembl_genes %>% filter(gene_biotype=="protein_coding") %>% nrow()
cat("Number of protein coding genes in original 71 cell types regions' data frame:\n")
enhancers %>% filter(!duplicated(TargetGene)) %>% filter(gene_biotype == "protein_coding") %>% nrow()
cat("Percentage of PCG in the regions data frame:\n")
paste0(100*round((enhancers %>% filter(!duplicated(TargetGene)) %>% filter(gene_biotype == "protein_coding") %>% nrow())/(ensembl_genes %>% filter(gene_biotype=="protein_coding") %>% nrow()), digits = 5), "%")
cat("Number of protein coding genes in selected 50 cell types regions' data frame:\n")
sel_enhancers %>% filter(!duplicated(TargetGene)) %>% filter(gene_biotype == "protein_coding") %>% nrow()
cat("Percentage of PCG in the selected regions data frame:\n")
paste0(100*round((sel_enhancers %>% filter(!duplicated(TargetGene)) %>% filter(gene_biotype == "protein_coding") %>% nrow())/(ensembl_genes %>% filter(gene_biotype=="protein_coding") %>% nrow()), digits = 5), "%")
```
## Save data

Save the selected regions

```{r}
saveRDS(object = sel_enhancers, file = paste0(in_dir, "/raw_enhancers/merged/05_raw_enhancers_selected_50cts_",date,".rds"))
```


























